---
title: "Airway RNA-seq: Dexamethasone vs Untreated (DESeq2)"
author: "Generated with Claude Code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
editor: source
---

```{r setup}
#| label: setup

# --- Packages -------------------------------------------------------------
# If any of these are missing, install once (uncomment and run in a console):
#
#   install.packages(c("tidyverse", "pheatmap", "RColorBrewer"))
#   if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
#   BiocManager::install(c("DESeq2", "apeglm", "EnhancedVolcano"))
#
# The document renders end-to-end once these are available.

suppressPackageStartupMessages({
  library(DESeq2)
  library(tidyverse)
  library(pheatmap)
  library(RColorBrewer)
})

# EnhancedVolcano is optional; fall back to a ggplot volcano if unavailable.
have_enhanced_volcano <- requireNamespace("EnhancedVolcano", quietly = TRUE)

# --- Paths ----------------------------------------------------------------
# Resolve the repository root robustly by searching upward from the working
# directory until we find the data/ directory. This keeps the document working
# whether it is rendered from the repo root or from a subfolder.
find_repo_root <- function(start = getwd(), marker = "data") {
  dir <- normalizePath(start, mustWork = FALSE)
  for (i in 1:6) {
    if (dir.exists(file.path(dir, marker)) &&
        file.exists(file.path(dir, marker, "experiment_table.csv"))) {
      return(dir)
    }
    parent <- dirname(dir)
    if (identical(parent, dir)) break
    dir <- parent
  }
  stop("Could not locate the repo root (a folder containing data/experiment_table.csv).")
}

here_root    <- find_repo_root()
counts_path  <- file.path(here_root, "data", "rsem.merged.gene_counts.tsv")
samples_path <- file.path(here_root, "data", "experiment_table.csv")

results_dir <- file.path(here_root, "results")
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

stopifnot(file.exists(counts_path), file.exists(samples_path))

# Analysis parameters
alpha_fdr        <- 0.05   # FDR threshold for calling DE genes
min_count        <- 10     # per-sample count floor for pre-filtering
min_samples      <- 4      # genes must clear min_count in >= this many samples
                           # (4 = size of the smallest treatment group)
```

# Overview

This document runs a standard bulk RNA-seq differential-expression analysis on the
**airway** dataset: primary human airway smooth muscle cells from four donors
(cell lines), each sequenced **Untreated** and treated with **Dexamethasone**.

The design is **paired** — the same donor appears in both conditions — so we block on
`cellLine` and test the `treatment` effect:

$$\text{design} = \sim \text{cellLine} + \text{treatment}$$

`Untreated` is the reference level, so a positive log2 fold-change means the gene is
**up in Dexamethasone** relative to Untreated.

# Load inputs

```{r load-inputs}
#| label: load-inputs

# Counts: gene_id, transcript_id(s), then one column per sample (SRA accession).
counts_raw <- readr::read_tsv(counts_path, show_col_types = FALSE)

# Sample sheet: ID (SRA accession), cellLine, treatment.
coldata_raw <- readr::read_csv(samples_path, show_col_types = FALSE)

dim(counts_raw)
coldata_raw
```

```{r build-matrix}
#| label: build-matrix

# Drop non-count columns and set Ensembl gene IDs as rownames.
count_mat <- counts_raw |>
  dplyr::select(-`transcript_id(s)`) |>
  tibble::column_to_rownames("gene_id") |>
  as.matrix()

# Sample sheet as a data frame with row order we control.
coldata <- as.data.frame(coldata_raw)
rownames(coldata) <- coldata$ID
```

# Align samples

The count columns must line up with the sample sheet rows exactly. We reorder the
matrix columns to match the sample-sheet `ID` order and then assert equality — if the
inputs ever change, this fails loudly instead of producing silently wrong results.

```{r align}
#| label: align

count_mat <- count_mat[, coldata$ID]
stopifnot(identical(colnames(count_mat), coldata$ID))

# Confirm the join is 1:1 over all 8 samples.
stopifnot(ncol(count_mat) == nrow(coldata))
colnames(count_mat)
```

# Round non-integer RSEM counts

RSEM assigns ambiguous reads probabilistically, so its gene-level counts are
**estimates, not integers** (e.g. `284.50`). DESeq2's negative-binomial model requires
integer counts, so we round before building the dataset. (The more principled
alternative is `tximport` on the per-sample RSEM files, which are not included in this
repository — only the merged gene-level table is.)

```{r round-counts}
#| label: round-counts

# Show that the raw matrix really is non-integer.
any_non_integer <- any(count_mat != round(count_mat))
cat("Matrix contains non-integer values:", any_non_integer, "\n")

count_mat <- round(count_mat)
storage.mode(count_mat) <- "integer"

cat("Rounded matrix dimensions:", nrow(count_mat), "genes x",
    ncol(count_mat), "samples\n")
```

# Factors and reference level

```{r factors}
#| label: factors

coldata$cellLine  <- factor(coldata$cellLine)
coldata$treatment <- factor(coldata$treatment)

# Untreated is the baseline; log2FC will read as Dexamethasone-vs-Untreated.
coldata$treatment <- relevel(coldata$treatment, ref = "Untreated")

levels(coldata$treatment)
table(coldata$cellLine, coldata$treatment)
```

# Build the DESeqDataSet and pre-filter

```{r build-dds}
#| label: build-dds

dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = coldata,
  design    = ~ cellLine + treatment
)

# Pre-filter: keep genes with at least `min_count` reads in at least
# `min_samples` samples. This removes the large block of all-zero / very low
# genes before fitting.
keep <- rowSums(counts(dds) >= min_count) >= min_samples
cat("Genes before filtering:", nrow(dds), "\n")
dds  <- dds[keep, ]
cat("Genes after filtering: ", nrow(dds), "\n")
```

# Run DESeq2

```{r run-deseq}
#| label: run-deseq

dds <- DESeq(dds)
resultsNames(dds)
```

# Quality control

Variance-stabilising transform, then PCA and a sample-distance heatmap. We expect
samples to separate by treatment while pairing by donor (cell line).

```{r vst}
#| label: vst

vsd <- vst(dds, blind = TRUE)
```

```{r pca}
#| label: fig-pca
#| fig-cap: "PCA of VST-transformed counts. Colour = treatment, shape = cell line."

pca_df <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
pct    <- round(100 * attr(pca_df, "percentVar"))

p_pca <- ggplot(pca_df, aes(PC1, PC2, colour = treatment, shape = cellLine)) +
  geom_point(size = 4) +
  labs(
    x = paste0("PC1: ", pct[1], "% variance"),
    y = paste0("PC2: ", pct[2], "% variance"),
    title = "PCA of VST counts"
  ) +
  theme_bw(base_size = 13)

p_pca
ggsave(file.path(results_dir, "pca.png"), p_pca, width = 7, height = 5, dpi = 150)
```

```{r sample-distance}
#| label: fig-sample-dist
#| fig-cap: "Euclidean distances between samples on VST-transformed counts."

sample_dists <- dist(t(assay(vsd)))
dist_mat <- as.matrix(sample_dists)
rownames(dist_mat) <- paste(vsd$treatment, vsd$cellLine, sep = " | ")
colnames(dist_mat) <- rownames(dist_mat)

pheatmap(
  dist_mat,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample-to-sample distances",
  filename = file.path(results_dir, "sample_distance_heatmap.png"),
  width = 7, height = 6
)
# Also draw inline (pheatmap writes to file when `filename` is set, so call again).
pheatmap(
  dist_mat,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample-to-sample distances"
)
```

# Differential expression results

```{r results}
#| label: results

res <- results(
  dds,
  contrast = c("treatment", "Dexamethasone", "Untreated"),
  alpha = alpha_fdr
)

# Shrunken log2 fold-changes for ranking and plotting.
res_shrunk <- lfcShrink(dds, coef = "treatment_Dexamethasone_vs_Untreated",
                        type = "apeglm")

summary(res)

n_de <- sum(res$padj < alpha_fdr, na.rm = TRUE)
cat("Genes with padj <", alpha_fdr, ":", n_de, "\n")
```

```{r results-table}
#| label: results-table

res_df <- as.data.frame(res) |>
  tibble::rownames_to_column("gene_id") |>
  dplyr::arrange(padj)

head(res_df, 15)
```

# Figures

```{r ma-plot}
#| label: fig-ma
#| fig-cap: "MA plot of shrunken log2 fold-changes (Dexamethasone vs Untreated)."

png(file.path(results_dir, "ma_plot.png"), width = 7, height = 5,
    units = "in", res = 150)
plotMA(res_shrunk, ylim = c(-5, 5), main = "MA plot (shrunken LFC)")
invisible(dev.off())

plotMA(res_shrunk, ylim = c(-5, 5), main = "MA plot (shrunken LFC)")
```

```{r volcano}
#| label: fig-volcano
#| fig-cap: "Volcano plot of differential expression."

if (have_enhanced_volcano) {
  p_volcano <- EnhancedVolcano::EnhancedVolcano(
    res_df,
    lab = res_df$gene_id,
    x = "log2FoldChange",
    y = "padj",
    pCutoff = alpha_fdr,
    FCcutoff = 1,
    title = "Dexamethasone vs Untreated",
    subtitle = NULL
  )
} else {
  # ggplot fallback so the document always renders.
  volc_df <- res_df |>
    dplyr::filter(!is.na(padj)) |>
    dplyr::mutate(
      sig = dplyr::case_when(
        padj < alpha_fdr & log2FoldChange >=  1 ~ "Up",
        padj < alpha_fdr & log2FoldChange <= -1 ~ "Down",
        TRUE ~ "NS"
      )
    )
  p_volcano <- ggplot(volc_df,
                      aes(log2FoldChange, -log10(padj), colour = sig)) +
    geom_point(alpha = 0.6, size = 1.4) +
    scale_colour_manual(values = c(Up = "firebrick", Down = "steelblue",
                                   NS = "grey70")) +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey50") +
    geom_hline(yintercept = -log10(alpha_fdr), linetype = "dashed",
               colour = "grey50") +
    labs(title = "Dexamethasone vs Untreated", colour = NULL) +
    theme_bw(base_size = 13)
}

p_volcano
ggsave(file.path(results_dir, "volcano.png"), p_volcano,
       width = 8, height = 6, dpi = 150)
```

```{r top-gene-heatmap}
#| label: fig-top-heatmap
#| fig-cap: "Top 30 DE genes (by adjusted p-value) on VST-transformed counts, row-scaled."

top_genes <- res_df |>
  dplyr::filter(!is.na(padj)) |>
  dplyr::slice_min(padj, n = 30, with_ties = FALSE) |>
  dplyr::pull(gene_id)

mat <- assay(vsd)[top_genes, , drop = FALSE]
mat <- t(scale(t(mat)))  # row-scale (z-score per gene)

annot_col <- as.data.frame(colData(vsd)[, c("treatment", "cellLine")])

pheatmap(
  mat,
  annotation_col = annot_col,
  show_colnames = FALSE,
  main = "Top 30 DE genes (VST, row-scaled)",
  filename = file.path(results_dir, "top_genes_heatmap.png"),
  width = 8, height = 8
)
pheatmap(
  mat,
  annotation_col = annot_col,
  show_colnames = FALSE,
  main = "Top 30 DE genes (VST, row-scaled)"
)
```

# Export results

```{r export}
#| label: export

out_csv <- file.path(results_dir, "deseq2_results.csv")
readr::write_csv(res_df, out_csv)
cat("Wrote full results table to:", out_csv, "\n")
cat("Figures written to:", results_dir, "\n")
```

# Session info

```{r session-info}
#| label: session-info

sessionInfo()
```
