---
title: "Dexamethasone response in airway smooth muscle"
subtitle: "Bulk RNA-seq differential expression with DESeq2, paired design `~ cellLine + treatment`"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    embed-resources: true
    fig-width: 7
    fig-height: 5
knitr:
  opts_chunk:
    message: false
    warning: false
    dev: ["png", "pdf"]
    fig.path: "figures/"
    dpi: 150
---

## Overview

This document runs the complete differential expression workflow for the **airway** dataset
(Himes et al. — human airway smooth muscle cells treated with dexamethasone) end to end:
input validation, count preparation, quality control, model fitting, and results export.

The experiment is a **paired design**. Four cell lines each contribute one `Untreated` and one
`Dexamethasone` sample, so cell line is a nuisance factor and not a replicate group. The model
blocks on it (`~ cellLine + treatment`, with `treatment` last so it is the coefficient of
interest). A diagnostic section near the end quantifies what an unpaired 4-vs-4 comparison would
cost, which is the single most likely way to get this analysis wrong.

Every consistency check in this document is a hard `stopifnot()` assertion rather than a printed
note. If an input file changes shape, a sample is mismatched, or the reference level flips, the
render fails loudly instead of producing plausible but wrong numbers.

```{r}
#| label: setup
set.seed(20260805)

suppressPackageStartupMessages({
  library(DESeq2)
  library(ggplot2)
  library(pheatmap)
  library(RColorBrewer)
})

dir.create("results", showWarnings = FALSE)
dir.create("figures", showWarnings = FALSE)

theme_set(theme_bw(base_size = 12))

# Diagnostics are printed with cat()/print() rather than message() so they always
# appear in the rendered document, independent of the chunk's message setting.
cat("DESeq2 version:", as.character(packageVersion("DESeq2")), "\n")
```

## Load and validate the inputs

`check.names = FALSE` is essential: it preserves the `transcript_id(s)` header, which contains
parentheses, and the sample names, instead of mangling them into `transcript_id.s.`. The counts
file has Windows line endings, which R handles transparently.

```{r}
#| label: load-inputs
coldata <- read.csv("data/experiment_table.csv", stringsAsFactors = FALSE)
raw     <- read.delim("data/rsem.merged.gene_counts.tsv", check.names = FALSE)

cat("sample sheet:  ", nrow(coldata), "rows x", ncol(coldata), "columns\n")
cat("counts file:   ", nrow(raw), "rows x", ncol(raw), "columns\n")
cat("counts columns:", paste(names(raw), collapse = ", "), "\n")
```

### Build the count matrix, matching samples by name

The first two columns are identifiers, not samples, and must be removed before a numeric matrix
can be built. The sample columns happen to be in the same order as the sample sheet rows, but
the matrix is reordered **by name** anyway — that is what keeps this document correct if the
counts file is ever regenerated with its columns in a different order. A silent mismatch here
would produce results that look entirely reasonable and are wrong.

```{r}
#| label: build-matrix
id_cols <- c("gene_id", "transcript_id(s)")
stopifnot(all(id_cols %in% names(raw)))
stopifnot(!anyDuplicated(raw$gene_id))

counts_df <- raw[, setdiff(names(raw), id_cols), drop = FALSE]
rownames(counts_df) <- raw$gene_id

# Fail rather than guess if the sample sets differ at all.
stopifnot(setequal(colnames(counts_df), coldata$ID))

counts_df <- counts_df[, coldata$ID, drop = FALSE]
stopifnot(identical(colnames(counts_df), coldata$ID))

# DESeq2 matches colData to countData by row name, so set them explicitly.
rownames(coldata) <- coldata$ID

cat("sample columns aligned to the sample sheet by name:",
    identical(colnames(counts_df), coldata$ID), "\n")
cat("samples:", ncol(counts_df), "\n")
```

### The design, made visible

Every cell of this cross-tab should contain exactly `1`. That is the paired design: each cell
line contributes one matched pair, so there are no within-group replicates to average over —
the treatment effect is estimated *within* cell line.

```{r}
#| label: design-table
design_tab <- table(cellLine = coldata$cellLine, treatment = coldata$treatment)
knitr::kable(design_tab, caption = "Samples per cell line and treatment.")

stopifnot(all(design_tab == 1))
cat("balanced paired design confirmed:", nrow(design_tab), "cell lines x",
    ncol(design_tab), "treatments, one sample each\n")
```

## Prepare counts for DESeq2

### Rounding

RSEM reports *expected* counts, which are fractional (e.g. `284.50`). DESeq2 requires integers
and refuses fractional input. The matrix is therefore rounded with `round()` — **not**
`as.integer()`, which truncates toward zero and would systematically bias every fractional
count downward. DESeq2 converts the rounded whole numbers to integer storage itself.

The usual route for RSEM output is `tximport`, which does not apply here: this file is already
merged to gene level by the nf-core/rnaseq pipeline.

```{r}
#| label: round-counts
counts_num <- as.matrix(counts_df)

n_nonint <- sum(apply(counts_num, 1, function(x) any(x != round(x))))
cat("gene rows containing at least one non-integer expected count:", n_nonint, "\n")

counts <- round(counts_num)
stopifnot(all(counts == round(counts)), !anyNA(counts))
```

### Factor levels and the reference

R orders factor levels alphabetically, which would make `Dexamethasone` the reference and
silently flip the sign of every log2 fold change. The reference is set explicitly, and asserted.

**With `Untreated` as the reference, a positive log2 fold change means the gene is induced by
dexamethasone.**

```{r}
#| label: factors
coldata$treatment <- relevel(factor(coldata$treatment), ref = "Untreated")
coldata$cellLine  <- factor(coldata$cellLine)

stopifnot(levels(coldata$treatment)[1] == "Untreated")

cat("treatment levels (reference first):",
    paste(levels(coldata$treatment), collapse = " -> "), "\n")
cat("cellLine levels:", paste(levels(coldata$cellLine), collapse = ", "), "\n")
```

### Build the object and pre-filter

The file is the full Ensembl annotation, so roughly half of it is zero in every sample.
Pre-filtering removes rows that cannot contribute evidence, which speeds up fitting and
improves the multiple-testing correction by shrinking the number of tests.

```{r}
#| label: build-dds
dds_all <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = coldata,
  design    = ~ cellLine + treatment
)

n_before  <- nrow(dds_all)
n_allzero <- sum(rowSums(counts(dds_all)) == 0)

keep <- rowSums(counts(dds_all)) >= 10
dds  <- dds_all[keep, ]

cat(sprintf("genes: %d in file -> %d retained (%.1f%%)\n",
            n_before, nrow(dds), 100 * nrow(dds) / n_before))
cat(sprintf("all-zero across all samples: %d (%.1f%%)\n",
            n_allzero, 100 * n_allzero / n_before))

# A parsing or filtering regression would show up immediately as a collapsed gene count.
stopifnot(nrow(dds) > 15000, ncol(dds) == nrow(coldata))
```

### Gene symbols (optional)

Gene IDs are unversioned Ensembl (`ENSG00000000003`), so they map directly with no `.N` suffix
to strip. Symbols make the tables and figures readable, but the mapping is optional: if
`org.Hs.eg.db` is not installed the document still renders, labelling everything with Ensembl
IDs instead. Unmapped IDs — common for non-coding entries — keep their Ensembl ID as the label.
The `gene_id` column is never replaced, only supplemented with `symbol`.

```{r}
#| label: annotate
have_org <- requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
            requireNamespace("AnnotationDbi", quietly = TRUE)

gene_map <- if (have_org) {
  suppressMessages(AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys      = rownames(dds),
    column    = "SYMBOL",
    keytype   = "ENSEMBL",
    multiVals = "first"
  ))
} else {
  setNames(rep(NA_character_, nrow(dds)), rownames(dds))
}

# Label with the symbol when there is one, otherwise fall back to the Ensembl ID.
gene_label <- function(ids) {
  s <- unname(gene_map[ids])
  ifelse(is.na(s) | s == "", ids, s)
}
symbol_of <- function(ids) unname(gene_map[ids])

# Reverse lookup, used only for the marker-gene positive control below.
id_for_symbol <- function(sym) {
  hits <- names(gene_map)[!is.na(gene_map) & gene_map == sym]
  if (length(hits) == 0) NA_character_ else hits[1]
}

if (have_org) {
  cat(sprintf("symbol mapping: %d of %d filtered genes mapped (%.1f%%)\n",
              sum(!is.na(gene_map)), length(gene_map),
              100 * mean(!is.na(gene_map))))
} else {
  cat("org.Hs.eg.db not installed - labelling with Ensembl IDs only.\n")
  cat("Install it with: BiocManager::install('org.Hs.eg.db')\n")
}
```

## Quality control

### Library sizes

Library sizes span roughly a two-fold range, which is exactly why comparisons must be made on
size-factor-normalised data rather than raw counts.

```{r}
#| label: fig-library-sizes
#| fig-cap: "Total expected counts per sample, after filtering."
lib <- data.frame(
  sample    = colnames(dds),
  total     = colSums(counts(dds)),
  cellLine  = dds$cellLine,
  treatment = dds$treatment
)

ggplot(lib, aes(x = reorder(sample, total), y = total / 1e6, fill = treatment)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Total counts (millions)", fill = "Treatment",
       title = "Library sizes") +
  scale_fill_brewer(palette = "Set2")

cat(sprintf("library sizes: %.1fM - %.1fM (%.2f-fold spread)\n",
            min(lib$total) / 1e6, max(lib$total) / 1e6,
            max(lib$total) / min(lib$total)))
```

### Variance-stabilising transform

`vst()` is used for visualisation and clustering only — the differential test itself runs on raw
counts. `blind = FALSE` lets the design inform the dispersion trend used by the transform, which
is the right choice for QC of an experiment whose design is already known.

```{r}
#| label: vst
vsd <- vst(dds, blind = FALSE)
cat("VST matrix:", nrow(vsd), "genes x", ncol(vsd), "samples\n")
```

### PCA

This is the visual argument for the blocked design. Expect the dominant component to separate
**cell lines**, with the treatment shift appearing as a small, consistent displacement *within*
each pair. The dashed lines connect the two samples of each cell line: if those connectors are
roughly parallel, the treatment effect is consistent across cell lines and blocking on cell
line will recover it.

```{r}
#| label: fig-pca
#| fig-cap: "PCA of VST-transformed counts. Dashed lines connect the paired samples of each cell line."
pca_data <- plotPCA(vsd, intgroup = c("cellLine", "treatment"), returnData = TRUE)
pct <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2)) +
  geom_line(aes(group = cellLine), colour = "grey60", linetype = 2) +
  geom_point(aes(colour = treatment, shape = cellLine), size = 4) +
  labs(x = paste0("PC1: ", pct[1], "% variance"),
       y = paste0("PC2: ", pct[2], "% variance"),
       colour = "Treatment", shape = "Cell line",
       title = "Samples cluster by cell line; treatment shifts within pairs") +
  scale_colour_brewer(palette = "Set1") +
  coord_fixed()
```

### Sample-to-sample distances

Same-cell-line pairs should be each other's nearest neighbours — another view of the same fact.

```{r}
#| label: fig-sample-distances
#| fig-cap: "Euclidean distances between samples on VST-transformed counts."
#| fig-height: 5.5
sample_dists <- dist(t(assay(vsd)))
dist_mat <- as.matrix(sample_dists)

sample_labels <- paste(vsd$cellLine, vsd$treatment, sep = " / ")
rownames(dist_mat) <- sample_labels
colnames(dist_mat) <- sample_labels

pheatmap(
  dist_mat,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  col  = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample distances (VST)"
)
```

## Fit the model

```{r}
#| label: fit
dds <- DESeq(dds)
print(resultsNames(dds))
```

The dispersion plot is the standard check that the fit is sane: per-gene estimates scattered
around a decreasing trend, with final values shrunk toward it.

```{r}
#| label: fig-dispersion
#| fig-cap: "Dispersion estimates: per-gene (black), fitted trend (red), shrunken final (blue)."
plotDispEsts(dds, main = "Dispersion estimates")
```

### Extract the contrast

The coefficient name is asserted before use, so the wrong contrast cannot be extracted quietly.

Two result objects are produced deliberately:

- **`res`** — unshrunken. Supplies the test statistics, p-values and adjusted p-values.
- **`resLFC`** — shrunken fold changes. Supplies the effect sizes used for ranking and plotting.

Shrinkage pulls in the wildly overstated fold changes that low-count genes produce, without
touching the p-values. Mixing the two up — for example filtering on shrunken effect sizes while
quoting unshrunken ones — is a common error, so both are carried side by side in the exported
table with explicit column names.

```{r}
#| label: results
coef_name <- "treatment_Dexamethasone_vs_Untreated"
stopifnot(coef_name %in% resultsNames(dds))

res <- results(dds, name = coef_name, alpha = 0.05)
summary(res)

# apeglm is the preferred shrinkage estimator; fall back to DESeq2's built-in
# "normal" so a missing optional package cannot break the render.
shrink_type <- if (requireNamespace("apeglm", quietly = TRUE)) "apeglm" else "normal"
cat("LFC shrinkage estimator:", shrink_type, "\n")

resLFC <- lfcShrink(dds, coef = coef_name, type = shrink_type)
```

```{r}
#| label: results-table
res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)

lfc_df <- as.data.frame(resLFC)
stopifnot(setequal(rownames(lfc_df), res_df$gene_id))

res_df$symbol                <- symbol_of(res_df$gene_id)
res_df$log2FoldChange_shrunk <- lfc_df[res_df$gene_id, "log2FoldChange"]
res_df$lfcSE_shrunk          <- lfc_df[res_df$gene_id, "lfcSE"]

res_df <- res_df[, c("gene_id", "symbol", "baseMean",
                     "log2FoldChange", "lfcSE",
                     "log2FoldChange_shrunk", "lfcSE_shrunk",
                     "stat", "pvalue", "padj")]
res_df <- res_df[order(res_df$padj, -abs(res_df$log2FoldChange_shrunk)), ]

n_tested <- sum(!is.na(res_df$padj))
n_sig    <- sum(res_df$padj < 0.05, na.rm = TRUE)
n_up     <- sum(res_df$padj < 0.05 & res_df$log2FoldChange_shrunk > 0, na.rm = TRUE)
n_down   <- sum(res_df$padj < 0.05 & res_df$log2FoldChange_shrunk < 0, na.rm = TRUE)
n_strong <- sum(res_df$padj < 0.05 & abs(res_df$log2FoldChange_shrunk) > 1, na.rm = TRUE)

cat(sprintf("genes tested (padj not NA): %d\n", n_tested))
cat(sprintf("padj < 0.05:                %d  (%d up, %d down)\n", n_sig, n_up, n_down))
cat(sprintf("padj < 0.05 & |LFC| > 1:    %d\n", n_strong))
```

### Top differentially expressed genes

```{r}
#| label: top-genes-table
top_tab <- head(res_df, 20)
top_tab$gene <- gene_label(top_tab$gene_id)
knitr::kable(
  top_tab[, c("gene", "gene_id", "baseMean", "log2FoldChange_shrunk", "pvalue", "padj")],
  digits = c(0, 0, 1, 3, 30, 30), row.names = FALSE,
  caption = "Top 20 genes by adjusted p-value. Positive log2FC = induced by dexamethasone."
)
```

## Figures

### MA plots

Shrunken and unshrunken side by side. The difference is concentrated at the low-expression left
edge, where the unshrunken estimates fan out into noise.

```{r}
#| label: fig-ma-plots
#| fig-cap: "MA plots. Left: shrunken fold changes. Right: unshrunken."
#| fig-width: 10
#| fig-height: 4.5
op <- par(mfrow = c(1, 2))
plotMA(resLFC, ylim = c(-5, 5), main = paste0("Shrunken (", shrink_type, ")"))
plotMA(res,    ylim = c(-5, 5), main = "Unshrunken")
par(op)
```

### Volcano

```{r}
#| label: fig-volcano
#| fig-cap: "Volcano plot using shrunken fold changes. Dashed lines mark padj = 0.05 and |log2FC| = 1."
#| fig-height: 6
vol <- res_df[!is.na(res_df$padj), ]
vol$category <- with(vol, ifelse(
  padj < 0.05 & abs(log2FoldChange_shrunk) > 1, "padj<0.05 & |LFC|>1",
  ifelse(padj < 0.05, "padj<0.05", "not significant")
))
vol$category <- factor(vol$category,
  levels = c("not significant", "padj<0.05", "padj<0.05 & |LFC|>1"))
vol$label <- gene_label(vol$gene_id)

top_lab <- head(vol[order(vol$padj), ], 15)

p_vol <- ggplot(vol, aes(x = log2FoldChange_shrunk, y = -log10(padj))) +
  geom_point(aes(colour = category), alpha = 0.6, size = 1.2) +
  geom_hline(yintercept = -log10(0.05), linetype = 2, colour = "grey40") +
  geom_vline(xintercept = c(-1, 1),     linetype = 2, colour = "grey40") +
  scale_colour_manual(values = c("not significant"      = "grey70",
                                 "padj<0.05"            = "#3182bd",
                                 "padj<0.05 & |LFC|>1"  = "#e6550d")) +
  labs(x = "log2 fold change (Dexamethasone vs Untreated, shrunken)",
       y = "-log10 adjusted p-value", colour = NULL,
       title = "Dexamethasone response") +
  theme(legend.position = "top")

if (requireNamespace("ggrepel", quietly = TRUE)) {
  p_vol <- p_vol + ggrepel::geom_text_repel(
    data = top_lab, aes(label = label),
    size = 3, max.overlaps = 20, min.segment.length = 0, show.legend = FALSE
  )
} else {
  p_vol <- p_vol + geom_text(
    data = top_lab, aes(label = label),
    size = 3, vjust = -0.7, check_overlap = TRUE, show.legend = FALSE
  )
}
p_vol
```

### p-value distribution

A well-behaved test gives a flat null with a spike near zero. A different shape (a slope, or a
spike at 1) points at an unmodelled batch effect or a mis-specified design.

```{r}
#| label: fig-pvalue-hist
#| fig-cap: "Raw p-value distribution for genes with a non-NA p-value."
#| fig-height: 4
ggplot(res_df[!is.na(res_df$pvalue), ], aes(x = pvalue)) +
  geom_histogram(binwidth = 0.02, fill = "#3182bd", colour = "white") +
  labs(x = "raw p-value", y = "genes", title = "p-value distribution")
```

### Top-gene heatmap

Row-centred VST values, so the colour shows each gene's deviation from its own mean. Because
samples are ordered with the annotation bars visible, the paired structure is readable directly:
the treatment split should hold within each cell line.

```{r}
#| label: fig-top-genes-heatmap
#| fig-cap: "Top 40 genes by adjusted p-value, VST values centred per gene."
#| fig-height: 8
top40 <- head(res_df$gene_id[!is.na(res_df$padj)], 40)

mat <- assay(vsd)[top40, , drop = FALSE]
mat <- mat - rowMeans(mat)
rownames(mat) <- make.unique(gene_label(top40))

ann_col <- data.frame(
  treatment = dds$treatment,
  cellLine  = dds$cellLine,
  row.names = paste(dds$cellLine, dds$treatment, sep = " / ")
)
colnames(mat) <- rownames(ann_col)

pheatmap(
  mat,
  annotation_col = ann_col,
  col = colorRampPalette(rev(brewer.pal(11, "RdBu")))(255),
  cluster_cols = TRUE, show_colnames = TRUE, fontsize_row = 8,
  main = "Top 40 DE genes (centred VST)"
)
```

### Marker genes: biological positive control

A handful of genes are well-established dexamethasone responders in airway smooth muscle. If the
analysis is correct these should show a clear, consistent shift, and it should be in the same
direction in all four cell lines. Genes are looked up **by symbol** through the annotation map
rather than by hardcoded Ensembl IDs; anything that fails to resolve or was removed by filtering
is reported and skipped rather than raising an error.

```{r}
#| label: fig-marker-genes
#| fig-cap: "Normalised counts for known dexamethasone-responsive genes. Lines connect paired samples."
#| fig-height: 6
markers <- c("CRISPLD2", "DUSP1", "PER1", "KLF15", "FKBP5", "ZBTB16", "SPARCL1")

if (!have_org) {
  cat("Annotation package unavailable - marker-gene control skipped.\n")
} else {
  marker_ids <- vapply(markers, id_for_symbol, character(1))
  usable     <- !is.na(marker_ids) & marker_ids %in% rownames(dds)

  if (any(!usable)) {
    cat("skipped (symbol unresolved or gene filtered out):",
        paste(markers[!usable], collapse = ", "), "\n")
  }

  if (!any(usable)) {
    cat("No marker genes available to plot.\n")
  } else {
    marker_counts <- do.call(rbind, lapply(markers[usable], function(sym) {
      d <- plotCounts(dds, gene = marker_ids[[sym]],
                      intgroup = c("cellLine", "treatment"),
                      returnData = TRUE, normalized = TRUE)
      d$symbol <- sym
      d
    }))

    print(
      ggplot(marker_counts, aes(x = treatment, y = count)) +
        geom_line(aes(group = cellLine), colour = "grey65", linetype = 2) +
        geom_point(aes(colour = cellLine), size = 2.5) +
        facet_wrap(~ symbol, scales = "free_y") +
        scale_y_log10() +
        labs(x = NULL, y = "normalised count (log10)", colour = "Cell line",
             title = "Known dexamethasone responders")
    )

    marker_stats <- res_df[res_df$gene_id %in% marker_ids[usable], ]
    marker_stats$gene <- gene_label(marker_stats$gene_id)
    knitr::kable(
      marker_stats[, c("gene", "gene_id", "baseMean",
                       "log2FoldChange_shrunk", "padj")],
      digits = c(0, 0, 1, 3, 30), row.names = FALSE,
      caption = "Marker gene statistics. Positive log2FC = induced by dexamethasone."
    )
  }
}
```

## Export results

```{r}
#| label: write-outputs
sig_df <- res_df[!is.na(res_df$padj) & res_df$padj < 0.05, ]

norm_counts <- counts(dds, normalized = TRUE)
norm_out <- data.frame(gene_id = rownames(norm_counts),
                       symbol  = symbol_of(rownames(norm_counts)),
                       norm_counts, check.names = FALSE)

vst_mat <- assay(vsd)
vst_out <- data.frame(gene_id = rownames(vst_mat),
                      symbol  = symbol_of(rownames(vst_mat)),
                      vst_mat, check.names = FALSE)

sf_out <- data.frame(sample       = names(sizeFactors(dds)),
                     size_factor  = as.numeric(sizeFactors(dds)),
                     library_size = colSums(counts(dds)),
                     cellLine     = as.character(dds$cellLine),
                     treatment    = as.character(dds$treatment))

write_tsv <- function(x, file) {
  write.table(x, file.path("results", file),
              sep = "\t", quote = FALSE, row.names = FALSE, na = "NA")
  cat(sprintf("  results/%-32s %6d rows\n", file, nrow(x)))
}

cat("written:\n")
write_tsv(res_df,      "de_results_all.tsv")
write_tsv(sig_df,      "de_results_significant.tsv")
write_tsv(norm_out,    "normalized_counts.tsv")
write_tsv(vst_out,     "vst_counts.tsv")
write_tsv(sf_out,      "size_factors.tsv")
```

## Diagnostic: what an unpaired analysis would cost

CLAUDE.md flags the unpaired 4-vs-4 comparison as the most likely mistake for this dataset. This
section makes the cost concrete by refitting the *same filtered counts* with `~ treatment`
alone, discarding the pairing.

This is a diagnostic only. **The blocked model above is the result**; the unpaired fit is shown
to justify why `cellLine` is in the formula.

```{r}
#| label: unpaired-fit
# Rebuilt from scratch rather than re-assigning design() on the fitted object,
# which would leave stale estimates attached.
dds_unpaired <- DESeqDataSetFromMatrix(
  countData = counts(dds),
  colData   = coldata,
  design    = ~ treatment
)
dds_unpaired <- DESeq(dds_unpaired)
stopifnot(coef_name %in% resultsNames(dds_unpaired))

res_unpaired <- results(dds_unpaired, name = coef_name, alpha = 0.05)

n_sig_unpaired <- sum(res_unpaired$padj < 0.05, na.rm = TRUE)

comparison <- data.frame(
  model            = c("~ cellLine + treatment (paired)", "~ treatment (unpaired)"),
  significant_genes = c(n_sig, n_sig_unpaired)
)
knitr::kable(comparison, caption = "Genes at padj < 0.05 under each model.")

cat(sprintf("the unpaired model recovers %d of %d genes (%.0f%%)\n",
            n_sig_unpaired, n_sig, 100 * n_sig_unpaired / max(n_sig, 1)))
```

```{r}
#| label: fig-paired-vs-unpaired
#| fig-cap: "Evidence per gene under both models. Points above the diagonal gain power from blocking on cell line."
#| fig-height: 5.5
cmp <- data.frame(
  gene_id  = rownames(res_unpaired),
  unpaired = -log10(res_unpaired$padj),
  paired   = -log10(res$padj[match(rownames(res_unpaired), rownames(res))])
)
cmp <- cmp[is.finite(cmp$paired) & is.finite(cmp$unpaired), ]

ggplot(cmp, aes(x = unpaired, y = paired)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_abline(slope = 1, intercept = 0, colour = "#e6550d") +
  geom_hline(yintercept = -log10(0.05), linetype = 2, colour = "grey40") +
  geom_vline(xintercept = -log10(0.05), linetype = 2, colour = "grey40") +
  labs(x = "-log10 padj, unpaired (~ treatment)",
       y = "-log10 padj, paired (~ cellLine + treatment)",
       title = "Blocking on cell line recovers evidence the unpaired model loses")
```

Cell line is the largest source of variance in this experiment, as the PCA showed. The unpaired
model leaves that variance in the residual, inflating the dispersion estimate for every gene and
weakening every test. Blocking on cell line moves it into the model, where it belongs.

## Reproducibility

```{r}
#| label: session-info
cat("results/\n"); print(list.files("results"))
cat("\nfigures/\n"); print(list.files("figures"))
cat("\n")
sessionInfo()
```
