---
title: "Airway RNA-seq: Dexamethasone vs Untreated (DESeq2)"
subtitle: "Paired differential-expression analysis (`~ cellLine + treatment`)"
author: "ClawBio demo"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  warning: false
  message: false
editor: source
---

## Overview

This report runs the full bulk RNA-seq differential-expression workflow for the
**airway** experiment: human airway smooth-muscle cell lines treated with
dexamethasone versus untreated controls. The design is **paired** — each of four
donor cell lines contributes one treated and one untreated sample — so we block on
`cellLine` and test the `treatment` effect with the design `~ cellLine + treatment`.

Inputs (read from `data/`):

- `experiment_table.csv` — sample sheet (`ID`, `cellLine`, `treatment`), 8 samples.
- `rsem.merged.gene_counts.tsv` — RSEM gene-level **expected counts** (non-integer),
  one column per sample.

Key conventions (see `CLAUDE.md`):

- `Untreated` is the **reference** level, so log2 fold changes are
  Dexamethasone-vs-Untreated.
- RSEM expected counts are **non-integer** and are **rounded** before DESeq2.
- Ensembl IDs are mapped to gene symbols via `org.Hs.eg.db`.

> **Dependencies.** This notebook needs the following packages. They are *not*
> installed automatically — uncomment and run once on a fresh machine:
>
> ```r
> # install.packages("BiocManager")
> # BiocManager::install(c("DESeq2", "apeglm", "org.Hs.eg.db"))
> # install.packages(c("tidyverse", "pheatmap", "RColorBrewer", "ggrepel"))
> ```

```{r setup}
suppressPackageStartupMessages({
  library(DESeq2)
  library(ggplot2)
  library(dplyr)
  library(tibble)
  library(pheatmap)
  library(RColorBrewer)
  library(ggrepel)
  library(org.Hs.eg.db)
})

set.seed(1)

# All outputs are written here.
results_dir <- "results"
dir.create(results_dir, showWarnings = FALSE)
```

## Load and validate inputs

We read the sample sheet and the count matrix, then **subset and order the count
columns by the sample sheet** — this makes the column/row correspondence explicit
and robust to any future reordering rather than assuming it.

```{r load-data}
# --- Sample sheet -----------------------------------------------------------
coldata <- read.csv("data/experiment_table.csv", stringsAsFactors = FALSE)

# Untreated as the reference level => coefficient is Dexamethasone vs Untreated.
coldata$treatment <- relevel(factor(coldata$treatment), ref = "Untreated")
coldata$cellLine  <- factor(coldata$cellLine)
rownames(coldata) <- coldata$ID

# --- Count matrix -----------------------------------------------------------
# check.names = FALSE keeps the literal "transcript_id(s)" / SRR column names.
counts_raw <- read.delim(
  "data/rsem.merged.gene_counts.tsv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

# Every sample-sheet ID must be present as a count column.
stopifnot(all(coldata$ID %in% colnames(counts_raw)))

# Build the numeric matrix, columns ordered exactly as the sample sheet.
count_mat <- as.matrix(counts_raw[, coldata$ID, drop = FALSE])
rownames(count_mat) <- counts_raw$gene_id
mode(count_mat) <- "numeric"

# RSEM expected counts are non-integer -> round for DESeq2 (needs integers).
count_mat <- round(count_mat)
storage.mode(count_mat) <- "integer"

# Sanity checks: alignment and no NAs introduced.
stopifnot(identical(colnames(count_mat), coldata$ID))
stopifnot(!anyNA(count_mat))

cat(sprintf("Genes: %d   Samples: %d\n", nrow(count_mat), ncol(count_mat)))
```

The design table:

```{r show-coldata}
coldata[, c("ID", "cellLine", "treatment")]
```

Confirm the paired layout — every cell line should have exactly one `Untreated`
and one `Dexamethasone` sample:

```{r check-balance}
table(coldata$cellLine, coldata$treatment)
```

## Build the DESeq2 dataset

```{r build-dds}
dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = coldata,
  design    = ~ cellLine + treatment
)
dds
```

### Pre-filter low-count genes

The smallest group has 4 samples, so we keep genes with at least **10 counts in at
least 4 samples**. This removes the large block of all-zero / near-zero genes before
testing.

```{r prefilter}
n_before <- nrow(dds)
keep <- rowSums(counts(dds) >= 10) >= 4
dds  <- dds[keep, ]
n_after <- nrow(dds)

cat(sprintf("Genes before filter: %d\nGenes after filter:  %d  (removed %d)\n",
            n_before, n_after, n_before - n_after))
```

## Run the differential-expression test

```{r run-deseq}
dds <- DESeq(dds)
resultsNames(dds)
```

We extract the treatment contrast and apply **apeglm** shrinkage for stable effect
sizes and ranking.

```{r extract-results}
coef_name <- "treatment_Dexamethasone_vs_Untreated"
stopifnot(coef_name %in% resultsNames(dds))

res    <- results(dds, name = coef_name, alpha = 0.05)
resLFC <- lfcShrink(dds, coef = coef_name, type = "apeglm")

summary(res)
```

Number of significant genes at FDR < 0.05:

```{r sig-counts}
sig <- !is.na(res$padj) & res$padj < 0.05
cat(sprintf("Significant (padj < 0.05): %d\n  up:   %d\n  down: %d\n",
            sum(sig),
            sum(sig & res$log2FoldChange > 0),
            sum(sig & res$log2FoldChange < 0)))
```

## Quality control

Variance-stabilised counts (`vst`) are used for all exploratory plots.

```{r vst}
vsd <- vst(dds, blind = FALSE)
```

### PCA

Colour = treatment, shape = cell line. In a well-behaved paired design, PC1 or PC2
should separate treated from untreated while cell lines stay identifiable.

```{r pca, fig.width=7, fig.height=5}
pca_df <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
pct <- round(100 * attr(pca_df, "percentVar"))

ggplot(pca_df, aes(PC1, PC2, colour = treatment, shape = cellLine)) +
  geom_point(size = 4) +
  labs(x = sprintf("PC1: %d%% variance", pct[1]),
       y = sprintf("PC2: %d%% variance", pct[2]),
       title = "PCA of variance-stabilised counts") +
  theme_bw()
```

### Sample-to-sample distances

```{r dist-heatmap, fig.width=7, fig.height=6}
sample_dists <- dist(t(assay(vsd)))
dist_mat <- as.matrix(sample_dists)
labels <- paste(vsd$cellLine, vsd$treatment, sep = " | ")
rownames(dist_mat) <- labels
colnames(dist_mat) <- labels

pheatmap(dist_mat,
         clustering_distance_rows = sample_dists,
         clustering_distance_cols = sample_dists,
         col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
         main = "Sample-to-sample distances")
```

## Annotate and assemble results

Map unversioned Ensembl gene IDs to gene symbols. `multiVals = "first"` resolves the
occasional 1-to-many mapping deterministically.

```{r annotate}
res_df <- as.data.frame(resLFC) |>
  rownames_to_column("gene_id")

# Bring in the unshrunken padj / stat as the significance basis.
res_df$padj_deseq <- res$padj[match(res_df$gene_id, rownames(res))]

res_df$symbol <- mapIds(
  org.Hs.eg.db,
  keys      = res_df$gene_id,
  column    = "SYMBOL",
  keytype   = "ENSEMBL",
  multiVals = "first"
)

res_df <- res_df |>
  dplyr::select(gene_id, symbol, baseMean, log2FoldChange, lfcSE, padj = padj_deseq) |>
  arrange(padj)

head(res_df, 20)
```

## Visualisation

### MA plot

```{r ma-plot, fig.width=7, fig.height=5}
plotMA(resLFC, ylim = c(-5, 5), main = "MA plot (apeglm-shrunk LFC)")
```

### Volcano plot

```{r volcano, fig.width=7, fig.height=6}
volc <- res_df |>
  filter(!is.na(padj)) |>
  mutate(
    neglog10p = -log10(padj),
    status = case_when(
      padj < 0.05 & log2FoldChange >=  1 ~ "Up",
      padj < 0.05 & log2FoldChange <= -1 ~ "Down",
      TRUE                               ~ "n.s."
    )
  )

top_label <- volc |> filter(status != "n.s.") |> arrange(padj) |> head(15)

ggplot(volc, aes(log2FoldChange, neglog10p, colour = status)) +
  geom_point(alpha = 0.6, size = 1.4) +
  scale_colour_manual(values = c(Up = "#c0392b", Down = "#2c6fbb", `n.s.` = "grey70")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey50") +
  geom_text_repel(data = top_label,
                  aes(label = ifelse(is.na(symbol), gene_id, symbol)),
                  size = 3, max.overlaps = 20, show.legend = FALSE) +
  labs(x = "log2 fold change (Dex vs Untreated)",
       y = "-log10 adjusted p-value",
       colour = NULL,
       title = "Volcano: Dexamethasone vs Untreated") +
  theme_bw()
```

### Heatmap of top differential genes

```{r top-heatmap, fig.width=7, fig.height=8}
top_genes <- res_df |> filter(!is.na(padj)) |> arrange(padj) |> head(40) |> pull(gene_id)

mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat)))                     # row-wise z-score
row_labels <- res_df$symbol[match(top_genes, res_df$gene_id)]
row_labels <- ifelse(is.na(row_labels), top_genes, row_labels)
rownames(mat) <- row_labels

ann_col <- as.data.frame(colData(vsd)[, c("treatment", "cellLine")])

pheatmap(mat,
         annotation_col = ann_col,
         show_colnames = FALSE,
         color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(255),
         main = "Top 40 DE genes (vst, row z-scored)")
```

## Export outputs

```{r export}
# Full results table (all tested genes).
write.csv(res_df,
          file.path(results_dir, "deseq2_results_all.csv"),
          row.names = FALSE)

# Significant subset.
sig_df <- res_df |> filter(!is.na(padj) & padj < 0.05)
write.csv(sig_df,
          file.path(results_dir, "deseq2_results_significant.csv"),
          row.names = FALSE)

# Normalised counts for downstream use.
norm_counts <- counts(dds, normalized = TRUE) |>
  as.data.frame() |>
  rownames_to_column("gene_id")
write.csv(norm_counts,
          file.path(results_dir, "normalized_counts.csv"),
          row.names = FALSE)

cat(sprintf("Wrote to %s/:\n  deseq2_results_all.csv (%d genes)\n  deseq2_results_significant.csv (%d genes)\n  normalized_counts.csv\n",
            results_dir, nrow(res_df), nrow(sig_df)))
```

## Reproducibility

```{r session-info}
sessionInfo()
```
