---
title: "Airway RNA-seq: Dexamethasone vs Untreated (DESeq2)"
subtitle: "Paired design, blocked on cell line"
author: "Catarina Nunes"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    df-print: paged
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
editor: source
---

## Overview

Differential expression for the **airway** experiment: primary human airway smooth
muscle cells treated with **dexamethasone** (a glucocorticoid) versus **untreated**
controls. The design is **paired** — each of 4 donor cell lines contributes one
untreated and one treated sample — so we block on `cellLine` and test `treatment`:

$$\text{design} = \sim\ \text{cellLine} + \text{treatment}$$

`Untreated` is the reference level, so a **positive log2 fold change means induced by
dexamethasone**.

Inputs (see `CLAUDE.md`):

- `data/rsem.merged.gene_counts.tsv` — RSEM merged gene-level **expected counts**
  (fractional; rounded to integers below), ~58.7k genes × 8 samples.
- `data/experiment_table.csv` — sample sheet (`ID`, `cellLine`, `treatment`).

All outputs are written to `results/`.

## Setup

Install any missing CRAN / Bioconductor packages, then load them.

```{r setup}
#| label: setup
pkgs_cran <- c("ggplot2", "pheatmap", "RColorBrewer", "ggrepel", "matrixStats")
pkgs_bioc <- c("DESeq2", "apeglm", "org.Hs.eg.db", "AnnotationDbi")

install_if_missing <- function(pkgs, bioc = FALSE) {
  missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing) == 0) return(invisible())
  if (bioc) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
    BiocManager::install(missing, update = FALSE, ask = FALSE)
  } else {
    install.packages(missing)
  }
}

install_if_missing(pkgs_cran, bioc = FALSE)
install_if_missing(pkgs_bioc, bioc = TRUE)

suppressPackageStartupMessages({
  library(DESeq2)
  library(ggplot2)
  library(pheatmap)
  library(RColorBrewer)
  library(ggrepel)
  library(apeglm)
  library(org.Hs.eg.db)
  library(AnnotationDbi)
})

# Output locations
dir.create("results", showWarnings = FALSE)
fig_dir <- file.path("results")

# Analysis parameters
ALPHA        <- 0.05   # FDR threshold
LFC_THRESH   <- 0      # log2FC threshold for calling DE (0 = any direction)
MIN_ROWSUM   <- 10     # pre-filter: keep genes with total counts >= this
N_TOP_HEATMAP <- 30    # genes shown in the DE heatmap

theme_set(theme_bw(base_size = 12))
```

## 1. Load inputs

```{r load}
#| label: load-data
counts_raw <- read.delim("data/rsem.merged.gene_counts.tsv",
                         check.names = FALSE, stringsAsFactors = FALSE)

# Drop non-count columns; gene IDs -> rownames
stopifnot(all(c("gene_id", "transcript_id(s)") %in% colnames(counts_raw)))
gene_ids <- counts_raw$gene_id
count_cols <- setdiff(colnames(counts_raw), c("gene_id", "transcript_id(s)"))
counts <- as.matrix(counts_raw[, count_cols])
rownames(counts) <- gene_ids

coldata <- read.csv("data/experiment_table.csv", stringsAsFactors = FALSE)

cat(sprintf("Counts matrix: %d genes x %d samples\n", nrow(counts), ncol(counts)))
knitr::kable(coldata, caption = "Sample sheet")
```

## 2. Round non-integer counts

RSEM reports **expected counts** (fractional). DESeq2 requires integers, so we round.

```{r round}
#| label: round-counts
frac_before <- mean(counts != round(counts))
counts <- round(counts)
mode(counts) <- "integer"
cat(sprintf("Rounded to integers (%.1f%% of cells were non-integer before rounding).\n",
            100 * frac_before))
```

## 3. Align samples to the sample sheet

Order the count columns to match sample-sheet rows and assert they agree — mismatched
columns would silently produce wrong results.

```{r align}
#| label: align-samples
stopifnot(all(coldata$ID %in% colnames(counts)))
counts <- counts[, coldata$ID]
stopifnot(identical(colnames(counts), coldata$ID))
cat("Count columns aligned to sample sheet.\n")
```

## 4. Factors & reference levels

`Untreated` is set as the reference so log2FC = Dexamethasone vs Untreated.

```{r factors}
#| label: factors
coldata$cellLine  <- factor(coldata$cellLine)
coldata$treatment <- relevel(factor(coldata$treatment), ref = "Untreated")

cat("treatment levels (first = reference):", paste(levels(coldata$treatment), collapse = ", "), "\n")
cat("cellLine  levels:", paste(levels(coldata$cellLine), collapse = ", "), "\n")
```

## 5. Build the DESeqDataSet & pre-filter

Paired/blocked design `~ cellLine + treatment`. We drop genes with very low total
counts (a large fraction are all-zero in this matrix).

```{r build}
#| label: build-dds
dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = coldata,
  design    = ~ cellLine + treatment
)

n_before <- nrow(dds)
keep <- rowSums(counts(dds)) >= MIN_ROWSUM
dds  <- dds[keep, ]
cat(sprintf("Pre-filter (rowSums >= %d): kept %d of %d genes (%.1f%%).\n",
            MIN_ROWSUM, nrow(dds), n_before, 100 * nrow(dds) / n_before))
```

## 6. Run DESeq2

```{r run}
#| label: run-deseq
dds <- DESeq(dds)
resultsNames(dds)
```

## 7. QC: variance-stabilised PCA & sample distances

We expect samples to **separate by treatment** and to **pair by cell line**.

```{r vst}
#| label: vst
vsd <- vst(dds, blind = TRUE)
```

### PCA

```{r pca}
#| label: fig-pca
#| fig-cap: "PCA on VST data. Colour = treatment, shape = donor cell line."
pca <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
pv  <- round(100 * attr(pca, "percentVar"))

p_pca <- ggplot(pca, aes(PC1, PC2, colour = treatment, shape = cellLine)) +
  geom_point(size = 4, alpha = 0.9) +
  labs(x = sprintf("PC1 (%d%%)", pv[1]),
       y = sprintf("PC2 (%d%%)", pv[2]),
       title = "PCA — VST counts") +
  scale_colour_brewer(palette = "Set1")
ggsave(file.path(fig_dir, "pca.png"), p_pca, width = 7, height = 5, dpi = 150)
p_pca
```

*Interpretation:* the dominant axis separating treated from untreated samples reflects
the dexamethasone response; consistent pairing by cell line motivates blocking on
`cellLine`.

### Sample-distance heatmap

```{r distheat}
#| label: fig-distheat
#| fig-cap: "Euclidean distances between samples on VST data."
sampleDists <- dist(t(assay(vsd)))
distMat <- as.matrix(sampleDists)
labs <- with(coldata, paste(treatment, cellLine, sep = " | "))
rownames(distMat) <- labs; colnames(distMat) <- labs
cols <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

png(file.path(fig_dir, "sample_distances.png"), width = 7, height = 6, units = "in", res = 150)
pheatmap(distMat, clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists, col = cols,
         main = "Sample-to-sample distances")
dev.off()
pheatmap(distMat, clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists, col = cols,
         main = "Sample-to-sample distances")
```

## 8. Results table

Extract the treatment contrast at FDR < `r ALPHA`, then apply `apeglm` LFC shrinkage
for ranking and plots.

```{r results}
#| label: deseq-results
res <- results(dds, name = "treatment_Dexamethasone_vs_Untreated",
               alpha = ALPHA)

res_shrunk <- lfcShrink(dds, coef = "treatment_Dexamethasone_vs_Untreated",
                        type = "apeglm", res = res)

summary(res)

n_sig  <- sum(res$padj < ALPHA, na.rm = TRUE)
n_up   <- sum(res$padj < ALPHA & res$log2FoldChange > 0, na.rm = TRUE)
n_down <- sum(res$padj < ALPHA & res$log2FoldChange < 0, na.rm = TRUE)
cat(sprintf("At FDR < %.2f: %d DE genes (%d up with dex, %d down).\n",
            ALPHA, n_sig, n_up, n_down))
```

## 9. Annotate with gene symbols

Map Ensembl gene IDs to HGNC symbols (`org.Hs.eg.db`).

```{r annotate}
#| label: annotate
res_df <- as.data.frame(res_shrunk)
res_df$gene_id <- rownames(res_df)

# Ensembl IDs here have no version suffix; map directly.
res_df$symbol <- mapIds(org.Hs.eg.db, keys = res_df$gene_id,
                        column = "SYMBOL", keytype = "ENSEMBL",
                        multiVals = "first")

res_df <- res_df[order(res_df$padj), ]
res_df <- res_df[, c("gene_id", "symbol", "baseMean", "log2FoldChange",
                     "lfcSE", "pvalue", "padj")]
knitr::kable(head(res_df, 15), digits = 3,
             caption = "Top 15 genes by adjusted p-value (shrunk log2FC)")
```

## 10. Figures

### MA plot

```{r ma}
#| label: fig-ma
#| fig-cap: "MA plot (apeglm-shrunk log2 fold changes). Red = FDR < alpha."
png(file.path(fig_dir, "ma_plot.png"), width = 7, height = 5, units = "in", res = 150)
plotMA(res_shrunk, alpha = ALPHA, main = "MA plot (shrunk LFC)")
dev.off()
plotMA(res_shrunk, alpha = ALPHA, main = "MA plot (shrunk LFC)")
```

### Volcano plot

```{r volcano}
#| label: fig-volcano
#| fig-cap: "Volcano plot. Labelled: top genes by adjusted p-value."
vol <- res_df
vol$sig <- with(vol, ifelse(!is.na(padj) & padj < ALPHA,
                            ifelse(log2FoldChange > 0, "Up", "Down"), "n.s."))
vol$neglog10p <- -log10(vol$pvalue)
top_lab <- head(vol[!is.na(vol$padj), ], 15)

p_vol <- ggplot(vol, aes(log2FoldChange, neglog10p, colour = sig)) +
  geom_point(alpha = 0.6, size = 1.4) +
  scale_colour_manual(values = c(Up = "#E41A1C", Down = "#377EB8", `n.s.` = "grey70")) +
  geom_text_repel(data = top_lab,
                  aes(label = ifelse(is.na(symbol), gene_id, symbol)),
                  size = 3, max.overlaps = 20, colour = "black") +
  labs(x = "log2 fold change (Dex vs Untreated)", y = "-log10 p-value",
       colour = NULL, title = "Volcano — Dexamethasone vs Untreated")
ggsave(file.path(fig_dir, "volcano.png"), p_vol, width = 7, height = 6, dpi = 150)
p_vol
```

### Heatmap of top DE genes

```{r topheat}
#| label: fig-topheat
#| fig-cap: "Top DE genes (VST, row z-scored)."
top_genes <- head(res_df$gene_id[!is.na(res_df$padj)], N_TOP_HEATMAP)
mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat)))  # z-score per gene

lab_symbol <- res_df$symbol[match(top_genes, res_df$gene_id)]
rownames(mat) <- ifelse(is.na(lab_symbol), top_genes, lab_symbol)
colnames(mat) <- with(coldata, paste(treatment, cellLine, sep = " | "))

ann_col <- data.frame(treatment = coldata$treatment, cellLine = coldata$cellLine)
rownames(ann_col) <- colnames(mat)

png(file.path(fig_dir, "top_de_heatmap.png"), width = 8, height = 9, units = "in", res = 150)
pheatmap(mat, annotation_col = ann_col, show_colnames = TRUE,
         main = sprintf("Top %d DE genes (z-scored VST)", length(top_genes)))
dev.off()
pheatmap(mat, annotation_col = ann_col, show_colnames = TRUE,
         main = sprintf("Top %d DE genes (z-scored VST)", length(top_genes)))
```

*Interpretation:* rows cluster into dexamethasone-induced vs -repressed genes, with
columns grouping by treatment across all four donors — the hallmark of a consistent,
donor-independent glucocorticoid response.

## 11. Export

```{r export}
#| label: export
out_csv <- file.path("results", "deseq2_results.csv")
write.csv(res_df, out_csv, row.names = FALSE)
cat("Wrote:\n  ", out_csv, "\n  ",
    paste(file.path(fig_dir, c("pca.png", "sample_distances.png", "ma_plot.png",
                               "volcano.png", "top_de_heatmap.png")),
          collapse = "\n   "), "\n")
```

## 12. Reproducibility

```{r sessioninfo}
#| label: sessioninfo
sessionInfo()
```
