---
title: "Bulk RNA-seq Differential Expression: Dexamethasone vs Untreated"
subtitle: "Airway smooth muscle cells, 4 donors, paired design (DESeq2)"
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
    df-print: paged
execute:
  warning: false
  message: false
---

## About this dataset

Airway smooth muscle cells from 4 donors (`cellLine`), each with one `Untreated` sample and
one `Dexamethasone`-treated sample (`treatment`) — the well-known "airway" dataset (Himes et
al.). Gene-level counts come from RSEM (`data/rsem.merged.gene_counts.tsv`); sample metadata
is in `data/experiment_table.csv`.

Because each donor contributes both conditions, this is a **paired design**: the model below
blocks on `cellLine` so that donor-to-donor variability doesn't obscure the treatment effect.

```{r}
#| label: setup
library(DESeq2)
library(apeglm)
library(dplyr)
library(readr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(AnnotationDbi)
library(org.Hs.eg.db)

# Thresholds for calling a gene "significant" — change these to re-run with different cutoffs.
PADJ_THRESHOLD <- 0.05
LFC_THRESHOLD <- 1
```

## Load and prepare data

```{r}
#| label: load-data
counts_raw <- read_tsv("data/rsem.merged.gene_counts.tsv", show_col_types = FALSE)

count_mat <- counts_raw |>
  dplyr::select(-`transcript_id(s)`) |>
  column_to_rownames("gene_id") |>
  as.matrix()

# RSEM reports non-integer "expected counts" (e.g. 340.80); DESeq2 requires integer counts.
count_mat <- round(count_mat)

coldata <- read_csv("data/experiment_table.csv", show_col_types = FALSE) |>
  column_to_rownames("ID")

# Align sample order and hard-fail if the ID join between the two files doesn't line up.
count_mat <- count_mat[, rownames(coldata)]
stopifnot(identical(colnames(count_mat), rownames(coldata)))

# "Untreated" as the reference level so fold changes read as Dexamethasone vs Untreated.
coldata$treatment <- factor(coldata$treatment, levels = c("Untreated", "Dexamethasone"))
coldata$cellLine <- factor(coldata$cellLine)

coldata
```

```{r}
#| label: prefilter
keep <- rowSums(count_mat) >= 10
cat(sprintf(
  "Keeping %d of %d genes after low-count filtering (rowSums >= 10).\n",
  sum(keep), length(keep)
))
count_mat <- count_mat[keep, ]
```

```{r}
#| label: build-dds
dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData = coldata,
  design = ~ cellLine + treatment
)
dds
```

## Quality control

### Library sizes

```{r}
#| label: qc-libsize
#| fig-cap: "Total counts per sample"
lib_sizes <- colSums(counts(dds)) |>
  enframe(name = "sample", value = "library_size") |>
  left_join(as.data.frame(colData(dds)) |> rownames_to_column("sample"), by = "sample")

ggplot(lib_sizes, aes(x = sample, y = library_size, fill = treatment)) +
  geom_col() +
  labs(title = "Library size per sample", x = NULL, y = "Total counts") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### PCA

Variance-stabilizing transformation (`vst`, blind to the design) followed by PCA. If the
design is doing what we expect, samples should separate more by donor (`cellLine`, driving
most of the variance between individuals) and/or by `treatment`.

```{r}
#| label: qc-pca
#| fig-cap: "PCA of VST-normalized counts"
vsd <- vst(dds, blind = TRUE)

pca_data <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment, shape = cellLine)) +
  geom_point(size = 4) +
  labs(
    x = paste0("PC1: ", percent_var[1], "% variance"),
    y = paste0("PC2: ", percent_var[2], "% variance"),
    title = "PCA of VST-normalized counts"
  ) +
  theme_minimal()
```

### Sample-to-sample distances

```{r}
#| label: qc-distance
#| fig-cap: "Sample-to-sample distances (VST)"
sample_dist <- dist(t(assay(vsd)))
dist_mat <- as.matrix(sample_dist)
rownames(dist_mat) <- paste(vsd$treatment, vsd$cellLine, sep = " - ")
colnames(dist_mat) <- NULL

colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
pheatmap(
  dist_mat,
  clustering_distance_rows = sample_dist,
  clustering_distance_cols = sample_dist,
  col = colors,
  main = "Sample-to-sample distances (VST)"
)
```

## Differential expression

```{r}
#| label: run-deseq
dds <- DESeq(dds)
plotDispEsts(dds, main = "Dispersion estimates")
```

```{r}
#| label: results
res <- results(dds, contrast = c("treatment", "Dexamethasone", "Untreated"), alpha = PADJ_THRESHOLD)
summary(res)
```

Shrink the log2 fold changes with `apeglm` — this pulls in noisy fold-change estimates for
low-count genes without touching genes with strong evidence, which is why the coefficient
below is used for reporting/plots rather than the raw Wald-test result above.

```{r}
#| label: shrink
stopifnot("treatment_Dexamethasone_vs_Untreated" %in% resultsNames(dds))
res_shrunk <- lfcShrink(dds, coef = "treatment_Dexamethasone_vs_Untreated", type = "apeglm")
```

### Annotate with gene symbols

The count matrix only has Ensembl gene IDs. We map them to gene symbols via `org.Hs.eg.db`,
keeping both columns in every output table — symbol for readability, Ensembl ID as the
unambiguous key (some IDs won't have a current symbol; those fall back to showing the
Ensembl ID).

```{r}
#| label: annotate
symbols <- mapIds(org.Hs.eg.db, keys = rownames(res_shrunk), column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")

res_df <- as.data.frame(res_shrunk) |>
  rownames_to_column("gene_id") |>
  as_tibble() |>
  mutate(
    symbol = symbols[gene_id],
    label = if_else(is.na(symbol), gene_id, symbol),
    significant = !is.na(padj) & padj < PADJ_THRESHOLD & abs(log2FoldChange) >= LFC_THRESHOLD
  ) |>
  relocate(symbol, .after = gene_id) |>
  arrange(padj)

n_sig <- sum(res_df$significant, na.rm = TRUE)
cat(sprintf(
  "%d genes significant at padj < %.2f and |log2FC| >= %.1f\n",
  n_sig, PADJ_THRESHOLD, LFC_THRESHOLD
))

res_df
```

### Sanity check against known biology

`FKBP5`, `TSC22D3`, `DUSP1`, and `PER1` are well-established glucocorticoid-responsive genes
in this dataset — they should show up here as strongly, significantly upregulated. If they
don't, something upstream (sample/ID alignment, factor levels, contrast direction) is wrong.

```{r}
#| label: sanity-check
res_df |>
  filter(symbol %in% c("FKBP5", "TSC22D3", "DUSP1", "PER1")) |>
  dplyr::select(gene_id, symbol, log2FoldChange, padj, significant)
```

## Visualization

### MA plot

```{r}
#| label: ma-plot
#| fig-cap: "MA plot (shrunken log2 fold changes)"
plotMA(res_shrunk, alpha = PADJ_THRESHOLD, main = "MA plot (shrunken LFC)")
```

### Volcano plot

```{r}
#| label: volcano
#| fig-cap: "Volcano plot: Dexamethasone vs Untreated"
plot_df <- res_df |> filter(!is.na(padj))
top_labels <- plot_df |> filter(significant) |> slice_min(padj, n = 10)

ggplot(plot_df, aes(x = log2FoldChange, y = -log10(padj), color = significant)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "grey70"), guide = "none") +
  geom_vline(xintercept = c(-LFC_THRESHOLD, LFC_THRESHOLD), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(PADJ_THRESHOLD), linetype = "dashed", color = "grey40") +
  geom_text_repel(data = top_labels, aes(label = label), size = 3, max.overlaps = 20) +
  labs(
    title = "Volcano plot: Dexamethasone vs Untreated",
    x = "log2 fold change (shrunken)", y = "-log10(adjusted p-value)"
  ) +
  theme_minimal()
```

### Heatmap of top differentially expressed genes

```{r}
#| label: heatmap-top-genes
#| fig-cap: "Top 30 differentially expressed genes (VST, row z-score)"
top_genes <- plot_df |> slice_min(padj, n = 30) |> pull(gene_id)

mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat)))
rownames(mat) <- res_df$label[match(top_genes, res_df$gene_id)]

annotation_col <- as.data.frame(colData(vsd)[, c("treatment", "cellLine")])

pheatmap(
  mat,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  main = "Top 30 differentially expressed genes (VST, row z-score)"
)
```

## Export results

```{r}
#| label: export
dir.create("results", showWarnings = FALSE)

write_csv(res_df, "results/de_results_all.csv")
write_csv(filter(res_df, significant), "results/de_results_significant.csv")

norm_counts <- counts(dds, normalized = TRUE) |>
  as.data.frame() |>
  rownames_to_column("gene_id")
write_csv(norm_counts, "results/normalized_counts.csv")

cat("Wrote results/de_results_all.csv, results/de_results_significant.csv, results/normalized_counts.csv\n")
```

## Session info

```{r}
#| label: session-info
sessioninfo::session_info()
```
