---
title: "Differential expression: Dexamethasone vs Untreated"
subtitle: "Human airway smooth muscle cells (Himes et al. 2014), paired design across 4 donors"
format:
  html:
    theme: [cosmo, custom.scss]
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
    df-print: kable
    tbl-cap-location: bottom
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(AnnotationDbi)
library(org.Hs.eg.db)

# Output directories, created up front so the render works from a clean checkout.
dir.create("results/tables",  recursive = TRUE, showWarnings = FALSE)
dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)

ALPHA <- 0.05  # FDR threshold used throughout

# ---- Purple-anchored palette (ColorBrewer Purples / PuOr; colourblind-safe) ----
# One purple/orange pair carries meaning across every figure:
#   purple = Dexamethasone / up-regulated, orange = Untreated / down-regulated.
PURPLE <- "#6a3d9a"
ORANGE <- "#e08214"
GREY   <- "#c4c0cc"

pal_treatment <- c(Untreated = ORANGE, Dexamethasone = PURPLE)
pal_direction <- c(
  "Up in Dexamethasone"   = PURPLE,
  "Down in Dexamethasone" = ORANGE,
  "Not significant"       = GREY
)

# Sequential purple ramp for distances; diverging orange<->purple for centred heatmaps.
seq_purples <- colorRampPalette(rev(RColorBrewer::brewer.pal(9, "Purples")))(255)
div_puor    <- colorRampPalette(RColorBrewer::brewer.pal(11, "PuOr"))(255)  # low=orange, high=purple

# Shared ggplot theme so PCA and volcano share one clean look.
theme_report <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "grey92"),
    plot.title    = element_text(face = "bold", colour = "#3f1f6b"),
    axis.title    = element_text(colour = "#26232b"),
    legend.position = "top",
    legend.title  = element_text(face = "bold")
  )
theme_set(theme_report)
```

## Experimental design

Human airway smooth muscle cells from four donor cell lines, each contributing one untreated and
one dexamethasone-treated sample — a fully balanced paired design. Dexamethasone is a
glucocorticoid, so the expected signal is a glucocorticoid receptor transcriptional response.

Because the same four donors appear in both treatment groups, `cellLine` belongs in the model as
a blocking factor. The design is therefore `~ cellLine + treatment`: donor-to-donor differences
are absorbed by the blocking term instead of inflating the residual variance, which is what makes
the paired design more powerful than an unpaired comparison of the same eight samples.

## Data loading

Two details in the counts file need care, both of which are guarded with assertions below rather
than left to trust:

- Column 2 is `transcript_id(s)`, a comma-separated transcript list — **not a sample**. Code that
  treats every column after the first as a count column silently produces a garbage ninth sample.
  It is dropped by name.
- `check.names = FALSE` is required when reading, because `make.names()` would rewrite the
  parentheses in `transcript_id(s)` and the drop-by-name would then miss.

```{r}
#| label: load-data
counts_raw <- read.delim(
  "data/rsem.merged.gene_counts.tsv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

coldata <- read.csv("data/experiment_table.csv", stringsAsFactors = FALSE)

# Drop the transcript list explicitly by name, never by position.
stopifnot("transcript_id(s)" %in% colnames(counts_raw))
counts_raw[["transcript_id(s)"]] <- NULL

# Gene IDs become the row key. They are bare Ensembl IDs with no version suffix and no
# duplicates, so they are safe to use directly.
stopifnot(
  !anyDuplicated(counts_raw$gene_id),
  all(grepl("^ENSG", counts_raw$gene_id))
)
# Named count_mat rather than `counts` to avoid shadowing DESeq2's counts() accessor.
count_mat <- as.matrix(counts_raw[, setdiff(colnames(counts_raw), "gene_id"), drop = FALSE])
rownames(count_mat) <- counts_raw$gene_id

# Exactly the 8 samples should remain once gene_id and the transcript list are gone.
stopifnot(ncol(count_mat) == nrow(coldata))

# Match samples by NAME, not by position. The two files happen to agree on order, but relying
# on that would break silently the moment either file is re-sorted.
stopifnot(setequal(colnames(count_mat), coldata$ID))
count_mat <- count_mat[, coldata$ID]
stopifnot(identical(colnames(count_mat), coldata$ID))

rownames(coldata) <- coldata$ID
```

Counts matrix: **`r format(nrow(count_mat), big.mark = ",")` genes × `r ncol(count_mat)` samples**,
with sample columns confirmed aligned to the metadata rows.

```{r}
#| label: design-table
coldata[, c("ID", "cellLine", "treatment")]
```

```{r}
#| label: design-balance
table(coldata$cellLine, coldata$treatment)
```

Each donor contributes exactly one sample per treatment — the design is balanced, with no
unpaired samples and no other covariates.

## Data characterisation

Three properties of RSEM output drive the preprocessing decisions that follow.

```{r}
#| label: characterisation
lib_sizes <- colSums(count_mat)

data.frame(
  Property = c(
    "Genes (rows)",
    "Genes zero in all 8 samples",
    "Genes with at least one fractional count",
    "Smallest library",
    "Largest library",
    "Library size ratio (max/min)"
  ),
  Value = c(
    format(nrow(count_mat), big.mark = ","),
    sprintf("%s (%.1f%%)",
            format(sum(rowSums(count_mat) == 0), big.mark = ","),
            100 * mean(rowSums(count_mat) == 0)),
    format(sum(apply(count_mat, 1, function(x) any(x != round(x)))), big.mark = ","),
    sprintf("%s (%s)", names(which.min(lib_sizes)),
            format(min(lib_sizes), big.mark = ",")),
    sprintf("%s (%s)", names(which.max(lib_sizes)),
            format(max(lib_sizes), big.mark = ",")),
    sprintf("%.2fx", max(lib_sizes) / min(lib_sizes))
  )
)
```

**Fractional counts.** RSEM apportions multi-mapping reads probabilistically, so values are
estimates rather than integers. DESeq2's negative binomial model requires integer counts, so the
matrix is rounded before the dataset is built.

**Near-half of all genes are zero everywhere.** These carry no information and only cost multiple
testing power, so they are removed by the low-count filter below.

**Library sizes span about two-fold.** Raw counts are therefore not comparable between samples;
every visual and distance-based comparison in this document uses normalised or
variance-stabilised values, never raw counts.

## Building the DESeqDataSet

The reference level is the one silent failure mode worth guarding hardest against. R orders factor
levels alphabetically, and `Dexamethasone` sorts before `Untreated` — so the *default* reference
is the treated group, which flips the sign of every log2 fold change without any error. After the
`relevel()` below, a positive log2FC means **up in Dexamethasone relative to Untreated**.

```{r}
#| label: build-dds
coldata$cellLine  <- factor(coldata$cellLine)
coldata$treatment <- relevel(factor(coldata$treatment), ref = "Untreated")

# Guard the sign convention: if this ever fails, every fold change in the report is backwards.
stopifnot(levels(coldata$treatment)[1] == "Untreated")

dds <- DESeqDataSetFromMatrix(
  countData = round(count_mat),       # RSEM estimates -> integers
  colData   = coldata,
  design    = ~ cellLine + treatment
)

n_before <- nrow(dds)
dds <- dds[rowSums(counts(dds)) >= 10, ]
n_after <- nrow(dds)
```

Filtering to genes with a total count of at least 10 across all samples takes
**`r format(n_before, big.mark = ",")` → `r format(n_after, big.mark = ",")` genes**
(`r sprintf("%.1f%%", 100 * (1 - n_after / n_before))` removed).

Note the order of operations here: counts are **rounded first, then filtered**, so the threshold
applies to the integers DESeq2 actually models. Filtering the raw fractional values instead would
leave 20,783 genes — 32 genes sit close enough to the boundary that rounding moves them across it.
The difference is immaterial to the results but worth knowing if this number is compared against
another pipeline.

```{r}
#| label: dataset-guards
# These guards are specific to THIS dataset. They exist so that swapping in different data fails
# loudly here rather than producing a plausible-looking but unverified report.
# Update the expected numbers deliberately if the input data changes.
stopifnot(n_before == 58735)

# Rounded-then-filtered count is 20,803. Kept as a tight range rather than an equality because
# the exact value depends on floating-point rounding of half-integer counts.
stopifnot(n_after > 20790, n_after < 20815)

resultsNames_expected <- "treatment_Dexamethasone_vs_Untreated"
```

## Quality control

QC uses variance-stabilising transformed (VST) counts. `blind = TRUE` means the transformation
ignores the design, so the clustering below is an independent check on the data rather than
something the model has already imposed on it.

```{r}
#| label: vst
vsd <- vst(dds, blind = TRUE)
```

### Principal component analysis

```{r}
#| label: fig-pca
#| fig-cap: "PCA of VST counts. Colour = treatment, shape = donor cell line."
pca_data <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
pct_var <- round(100 * attr(pca_data, "percentVar"))

p_pca <- ggplot(pca_data, aes(PC1, PC2, colour = treatment, shape = cellLine)) +
  geom_point(size = 4.5, alpha = 0.9, stroke = 1.1) +
  scale_colour_manual(values = pal_treatment) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(
    title = "PCA of variance-stabilised counts",
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance"),
    colour = "Treatment", shape = "Cell line"
  ) +
  # Two legends don't fit across the top of a 7-inch figure; stack them at the right.
  theme(legend.position = "right", legend.box = "vertical")

p_pca
ggsave("results/figures/pca.png", p_pca, width = 7, height = 5, dpi = 150)
```

Treatment is the dominant source of variation here: PC1 (42%) cleanly separates the four
untreated samples on the left from the four treated samples on the right, and donor identity
appears as the secondary PC2 (25%) spread within each treatment group. That donor spread on PC2
is real structure, and leaving it unmodelled would push it into the residuals and cost power —
which is exactly what the `cellLine` blocking term absorbs. So blocking on donor remains the
right call even though, in this dataset, the treatment effect is the larger of the two.

### Sample-to-sample distances

```{r}
#| label: fig-sample-dist
#| fig-cap: "Euclidean distances between samples on VST counts."
#| fig-height: 6
sample_dists <- dist(t(assay(vsd)))
dist_mat <- as.matrix(sample_dists)
rownames(dist_mat) <- paste(vsd$cellLine, vsd$treatment, sep = " / ")
colnames(dist_mat) <- colnames(vsd)

dist_args <- list(
  mat = dist_mat,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  col = seq_purples,   # dark purple = close, pale = distant
  border_color = "white",
  main = "Sample-to-sample distances (VST)"
)

do.call(pheatmap, dist_args)
do.call(pheatmap, c(dist_args, list(
  filename = "results/figures/sample_distances.png", width = 7, height = 6
)))
```

The dendrogram splits first into two clean blocks of four — all untreated on one side, all
treated on the other — reinforcing the PCA reading that treatment is the primary axis. Within
each block the donor pairing is only partial (N080611 is the one donor whose two samples sit
closest together), so the paired structure is present but secondary. Blocking on `cellLine`
still pays off by removing that residual donor variation from the treatment comparison.

## Differential expression

```{r}
#| label: run-deseq
dds <- DESeq(dds)
stopifnot(resultsNames_expected %in% resultsNames(dds))
```

### Dispersion estimates

```{r}
#| label: fig-dispersion
#| fig-cap: "Per-gene dispersion estimates, with the fitted trend and shrunken final values."
plotDispEsts(dds, main = "Dispersion estimates",
             genecol = GREY, fitcol = ORANGE, finalcol = PURPLE)

png("results/figures/dispersion.png", width = 7, height = 5, units = "in", res = 150)
plotDispEsts(dds, main = "Dispersion estimates",
             genecol = GREY, fitcol = ORANGE, finalcol = PURPLE)
invisible(dev.off())
```

A healthy plot shows dispersion falling as mean expression rises: the per-gene estimates (grey)
are shrunk toward the fitted trend (orange) to give the final values (purple). Genes whose
dispersion sits far above the trend are flagged as outliers and left unshrunken.

### Results

Two versions of the fold changes are computed and both are kept:

- **Unshrunken** (`res`) — the raw maximum-likelihood estimates, and the source of all *p*-values
  and adjusted *p*-values.
- **Shrunken** (`res_shrunk`, via `apeglm`) — fold changes moderated toward zero in proportion to
  how noisily they are measured. Low-count genes can otherwise show enormous but meaningless fold
  changes. These are the effect sizes reported in tables and used in every figure.

The *p*-values are identical between the two; only the effect sizes differ. Gene ranking
throughout is by adjusted *p*-value, not by fold change.

```{r}
#| label: results
res <- results(
  dds,
  contrast = c("treatment", "Dexamethasone", "Untreated"),
  alpha = ALPHA
)

res_shrunk <- lfcShrink(dds, coef = resultsNames_expected, type = "apeglm")

# apeglm returns rows in the same order; assert rather than assume before combining.
stopifnot(identical(rownames(res), rownames(res_shrunk)))
```

```{r}
#| label: results-summary
summary(res)
```

```{r}
#| label: results-counts
n_tested <- sum(!is.na(res$padj))
n_sig    <- sum(res$padj < ALPHA, na.rm = TRUE)
n_up     <- sum(res$padj < ALPHA & res$log2FoldChange > 0, na.rm = TRUE)
n_down   <- sum(res$padj < ALPHA & res$log2FoldChange < 0, na.rm = TRUE)
n_sig_lfc1 <- sum(res$padj < ALPHA & abs(res$log2FoldChange) > 1, na.rm = TRUE)

data.frame(
  Metric = c(
    "Genes tested (non-NA padj)",
    sprintf("Significant (padj < %.2f)", ALPHA),
    "  ... up in Dexamethasone",
    "  ... down in Dexamethasone",
    "Significant and |log2FC| > 1"
  ),
  Count = format(c(n_tested, n_sig, n_up, n_down, n_sig_lfc1), big.mark = ",")
)
```

Genes with `NA` adjusted *p*-values were removed by DESeq2's independent filtering — they are too
weakly expressed to have any realistic chance of detection, and excluding them raises the power
available to the remaining genes.

## Gene symbol annotation

The counts file carries bare Ensembl IDs. Symbols come from `org.Hs.eg.db`, an offline annotation
database, so the render needs no network and the annotation version is fixed.

The mapping is not one-to-one, and both failure modes are reported rather than silently absorbed.
The Ensembl ID stays the row key throughout; `symbol` is only ever a label. No row is dropped for
failing to map.

```{r}
#| label: annotate
symbols <- mapIds(
  org.Hs.eg.db,
  keys      = rownames(res),
  keytype   = "ENSEMBL",
  column    = "SYMBOL",
  multiVals = "first"   # keep one symbol per gene; collisions are counted below
)

res_df <- as.data.frame(res)
res_df$gene_id <- rownames(res_df)
res_df$symbol  <- unname(symbols[res_df$gene_id])
res_df$log2FoldChange_shrunk <- res_shrunk$log2FoldChange
res_df$lfcSE_shrunk          <- res_shrunk$lfcSE

res_df <- res_df[, c(
  "gene_id", "symbol", "baseMean",
  "log2FoldChange", "lfcSE", "log2FoldChange_shrunk", "lfcSE_shrunk",
  "stat", "pvalue", "padj"
)]
res_df <- res_df[order(res_df$padj, na.last = TRUE), ]

n_unmapped <- sum(is.na(res_df$symbol))
mapped_symbols <- res_df$symbol[!is.na(res_df$symbol)]
n_dup_symbols  <- sum(duplicated(mapped_symbols))

data.frame(
  Metric = c(
    "Genes in results table",
    "Ensembl IDs with no symbol",
    "Symbols shared by >1 Ensembl ID (duplicate rows)"
  ),
  Count = format(c(nrow(res_df), n_unmapped, n_dup_symbols), big.mark = ",")
)
```

Unmapped IDs are mostly retired Ensembl entries and non-coding genes with no assigned symbol.
Where several Ensembl IDs share a symbol, each remains its own row — collapsing them would merge
genuinely distinct features. This is why downstream work should join on `gene_id`, not `symbol`.

## Direction sanity check

The strongest available check that the sign convention is correct: three genes with well
established responses to glucocorticoid treatment. `DUSP1` and `KLF15` are induced by
dexamethasone; `CCL2` is repressed. If the reference level had defaulted to `Dexamethasone`, all
three signs would be inverted and the assertion below fails the render immediately — which is the
point of having it.

```{r}
#| label: sanity-check
marker_genes <- c("DUSP1", "KLF15", "CCL2")
expected_dir <- c(DUSP1 = 1, KLF15 = 1, CCL2 = -1)

marker_rows <- res_df[match(marker_genes, res_df$symbol), ]
stopifnot(!any(is.na(marker_rows$gene_id)))

marker_table <- data.frame(
  Symbol   = marker_rows$symbol,
  Gene     = marker_rows$gene_id,
  Expected = ifelse(expected_dir[marker_rows$symbol] > 0, "up", "down"),
  log2FC   = round(marker_rows$log2FoldChange_shrunk, 2),
  padj     = signif(marker_rows$padj, 3),
  Observed = ifelse(marker_rows$log2FoldChange_shrunk > 0, "up", "down")
)
marker_table
```

```{r}
#| label: sanity-assert
# Fail the render if any marker moves the wrong way.
stopifnot(all(
  sign(marker_rows$log2FoldChange_shrunk) == expected_dir[marker_rows$symbol]
))
```

All three markers move in the expected direction, so a positive log2FC in this report means up in
Dexamethasone relative to Untreated. The check here is on the *sign*, not significance: `DUSP1`
and `KLF15` are strongly induced with tiny adjusted *p*-values, while `CCL2` is repressed but sits
just above the FDR threshold (padj ≈ 0.05) — a weak effect in this dataset, though clearly in the
repressed direction.

## Figures

### MA plot

```{r}
#| label: fig-ma
#| fig-cap: "MA plot on shrunken fold changes. Purple points pass the FDR threshold."
plotMA(res_shrunk, alpha = ALPHA, ylim = c(-6, 6),
       colNonSig = GREY, colSig = PURPLE, colLine = ORANGE,
       main = "MA plot (apeglm-shrunken log2 fold changes)")

png("results/figures/ma_plot.png", width = 7, height = 5, units = "in", res = 150)
plotMA(res_shrunk, alpha = ALPHA, ylim = c(-6, 6),
       colNonSig = GREY, colSig = PURPLE, colLine = ORANGE,
       main = "MA plot (apeglm-shrunken log2 fold changes)")
invisible(dev.off())
```

Fold change against mean expression. Shrinkage pulls in the low-expression genes on the left that
would otherwise fan out into large, poorly-supported fold changes, leaving the significant calls
concentrated among adequately-expressed genes.

### Volcano plot

```{r}
#| label: fig-volcano
#| fig-cap: "Volcano plot on shrunken fold changes; the 15 genes with smallest padj are labelled."
#| fig-height: 6
volcano_df <- res_df[!is.na(res_df$padj), ]

# A padj that underflows to exactly 0 would plot at -log10(0) = Inf and wreck the y-axis.
# Floor it at the smallest non-zero value instead, and note if this ever applies.
min_nonzero_padj <- min(volcano_df$padj[volcano_df$padj > 0])
n_floored <- sum(volcano_df$padj == 0)
volcano_df$padj_plot <- pmax(volcano_df$padj, min_nonzero_padj)

volcano_df$significant <- with(
  volcano_df,
  ifelse(padj >= ALPHA, "Not significant",
         ifelse(log2FoldChange_shrunk > 0, "Up in Dexamethasone", "Down in Dexamethasone"))
)
volcano_df$label <- ifelse(is.na(volcano_df$symbol), volcano_df$gene_id, volcano_df$symbol)

label_df <- head(volcano_df[volcano_df$padj < ALPHA, ], 15)

# Draw significant points on top of the grey background so they read clearly.
volcano_df$significant <- factor(
  volcano_df$significant,
  levels = c("Not significant", "Down in Dexamethasone", "Up in Dexamethasone")
)
volcano_df <- volcano_df[order(volcano_df$significant), ]

p_volcano <- ggplot(volcano_df, aes(log2FoldChange_shrunk, -log10(padj_plot))) +
  geom_point(aes(colour = significant), size = 1.4, alpha = 0.7) +
  geom_hline(yintercept = -log10(ALPHA), linetype = "dashed", colour = "grey45") +
  geom_text_repel(
    data = label_df, aes(label = label),
    size = 3.2, fontface = "italic", colour = "#26232b",
    max.overlaps = Inf, min.segment.length = 0, box.padding = 0.4,
    segment.colour = "grey60"
  ) +
  scale_colour_manual(values = pal_direction, breaks = names(pal_direction)) +
  labs(
    title = "Dexamethasone vs Untreated",
    x = "log2 fold change (shrunken)",
    y = expression(-log[10]~adjusted~italic(p)),
    colour = NULL
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)))

p_volcano
ggsave("results/figures/volcano.png", p_volcano, width = 7, height = 6, dpi = 150)
```

The dashed line marks the FDR threshold. The response is bidirectional, as expected for a
glucocorticoid receptor programme that both induces and represses target genes.
`r if (n_floored > 0) paste0("Note: ", n_floored, " gene(s) had an adjusted p-value underflowing to zero and were floored to the smallest non-zero value for plotting only; the exported tables keep the raw values.") else ""`

### Heatmap of top differentially expressed genes

```{r}
#| label: fig-heatmap
#| fig-cap: "Top 30 genes by adjusted p-value, VST values centred per gene."
#| fig-height: 8
top_ids <- head(res_df$gene_id[!is.na(res_df$padj)], 30)

mat <- assay(vsd)[top_ids, ]
mat <- mat - rowMeans(mat)   # centre per gene, so colour shows relative change

top_labels <- res_df$symbol[match(top_ids, res_df$gene_id)]
# make.unique guards against two of the top 30 sharing a symbol, which would give the
# heatmap duplicate row labels.
rownames(mat) <- make.unique(ifelse(is.na(top_labels), top_ids, top_labels))

annotation_col <- data.frame(
  Treatment = colData(vsd)$treatment,
  CellLine  = colData(vsd)$cellLine,
  row.names = colnames(mat)
)

# Treatment colours match the rest of the report; donors get a light purple ramp.
cellline_levels <- levels(colData(vsd)$cellLine)
cellline_cols <- setNames(
  colorRampPalette(brewer.pal(9, "Purples")[4:8])(length(cellline_levels)),
  cellline_levels
)
annotation_colors <- list(
  Treatment = c(Untreated = ORANGE, Dexamethasone = PURPLE),
  CellLine  = cellline_cols
)

# Symmetric breaks so 0 sits on the neutral midpoint of the diverging ramp.
max_abs <- max(abs(mat))
breaks <- seq(-max_abs, max_abs, length.out = 256)

heatmap_args <- list(
  mat = mat,
  color = div_puor,               # orange = down, purple = up
  breaks = breaks,
  annotation_col = annotation_col,
  annotation_colors = annotation_colors,
  border_color = NA,
  show_colnames = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 8,
  main = "Top 30 DE genes (VST, gene-centred)"
)

do.call(pheatmap, heatmap_args)
do.call(pheatmap, c(heatmap_args, list(
  filename = "results/figures/top_genes_heatmap.png",
  width = 7, height = 8
)))
```

The columns split cleanly into an untreated block (left) and a treated block (right), matching
the treatment annotation track. The block of genes in the lower two-thirds — including DUSP1,
KLF15, SERPINA3, and SAMHD1 — is induced by dexamethasone, while the upper block (VCAM1, ADAM12,
CXCL12) is repressed. The donor annotation track shows no consistent ordering within either
block, consistent with the QC plots: treatment is the dominant signal and donor is the secondary
structure the `cellLine` term absorbs.

## Exported files

```{r}
#| label: exports
res_sig <- res_df[!is.na(res_df$padj) & res_df$padj < ALPHA, ]

norm_counts <- counts(dds, normalized = TRUE)
vst_counts  <- assay(vsd)

write.csv(res_df,  "results/tables/de_results_all.csv", row.names = FALSE)
write.csv(res_sig, "results/tables/de_results_significant.csv", row.names = FALSE)
write.csv(
  data.frame(gene_id = rownames(norm_counts), norm_counts, check.names = FALSE),
  "results/tables/normalized_counts.csv", row.names = FALSE
)
write.csv(
  data.frame(gene_id = rownames(vst_counts), vst_counts, check.names = FALSE),
  "results/tables/vst_counts.csv", row.names = FALSE
)

data.frame(
  File = c(
    "results/tables/de_results_all.csv",
    "results/tables/de_results_significant.csv",
    "results/tables/normalized_counts.csv",
    "results/tables/vst_counts.csv",
    "results/figures/pca.png",
    "results/figures/sample_distances.png",
    "results/figures/dispersion.png",
    "results/figures/ma_plot.png",
    "results/figures/volcano.png",
    "results/figures/top_genes_heatmap.png"
  ),
  Contents = c(
    sprintf("All %s tested genes, ordered by padj", format(nrow(res_df), big.mark = ",")),
    sprintf("%s genes at padj < %.2f", format(nrow(res_sig), big.mark = ","), ALPHA),
    "Median-of-ratios normalised counts",
    "Variance-stabilised counts",
    "PCA", "Sample distance heatmap", "Dispersion estimates",
    "MA plot", "Volcano plot", "Top 30 DE gene heatmap"
  )
)
```

## Top 20 differentially expressed genes

```{r}
#| label: top-table
top20 <- head(res_sig, 20)
data.frame(
  Symbol = ifelse(is.na(top20$symbol), "-", top20$symbol),
  Gene   = top20$gene_id,
  baseMean = round(top20$baseMean),
  log2FC = round(top20$log2FoldChange_shrunk, 2),
  padj   = signif(top20$padj, 3)
)
```

## Summary

Dexamethasone treatment of human airway smooth muscle cells produced
**`r format(n_sig, big.mark = ",")` differentially expressed genes** at an FDR of
`r ALPHA` out of `r format(n_tested, big.mark = ",")` tested
(`r format(n_up, big.mark = ",")` up, `r format(n_down, big.mark = ",")` down), using a paired
design that blocks on donor cell line. The direction of the response was confirmed against three
known glucocorticoid-responsive genes before any results were interpreted.

## Session information

```{r}
#| label: session-info
sessionInfo()
```
