---
title: "Dexamethasone Differential Gene Expression (airway RNA-seq)"
subtitle: "DESeq2 workflow — paired design across 4 airway smooth muscle cell lines"
author: "Generated with Claude Code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: paged
    fig-width: 7
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
editor: source
---

# Overview

This document runs a complete bulk RNA-seq differential gene expression (DGE)
analysis of the classic **airway** dataset (Himes et al. 2014, PMID: 24926665):
the effect of **dexamethasone** (a glucocorticoid) on gene expression in human
airway smooth muscle cells.

**Design.** Four cell lines (`N61311`, `N052611`, `N080611`, `N061011`) were each
profiled Untreated and Dexamethasone-treated — a **paired** design. Cell line is
therefore modelled as a blocking factor:

$$\text{design} = \sim \text{cellLine} + \text{treatment}$$

so that the `treatment` effect is estimated *within* cell line, controlling for
baseline between-line differences.

Outputs are written to `results/` (tables) and `figures/` (plots).

# Setup

```{r}
#| label: setup

# --- One-time install (uncomment on a clean machine) -----------------------
# install.packages(c("BiocManager", "ggplot2", "ggrepel", "pheatmap",
#                     "RColorBrewer", "readr", "dplyr", "tibble"))
# BiocManager::install(c("DESeq2", "apeglm"))
# # Optional gene-symbol annotation:
# BiocManager::install("org.Hs.eg.db")

suppressPackageStartupMessages({
  library(DESeq2)
  library(readr)
  library(dplyr)
  library(tibble)
  library(ggplot2)
  library(ggrepel)
  library(pheatmap)
  library(RColorBrewer)
})

set.seed(1)

# Output directories
dir.create("results", showWarnings = FALSE)
dir.create("figures", showWarnings = FALSE)

# Analysis thresholds
alpha_padj <- 0.05   # adjusted p-value significance cutoff
lfc_thresh <- 1      # |log2FC| threshold for volcano highlighting
```

# Load and sanitise data

We read the RSEM gene-level count matrix, drop the `transcript_id(s)` column,
and **round counts to integers** — RSEM emits fractional (expected) counts, but
DESeq2's negative-binomial model requires integers.

```{r}
#| label: load-data

# --- Count matrix ----------------------------------------------------------
raw <- read_tsv("data/rsem.merged.gene_counts.tsv", show_col_types = FALSE)

count_mat <- raw |>
  select(-`transcript_id(s)`) |>
  column_to_rownames("gene_id") |>
  as.matrix()

# RSEM expected counts are floats -> round to integer for DESeq2
count_mat <- round(count_mat)
storage.mode(count_mat) <- "integer"

# --- Sample metadata -------------------------------------------------------
coldata <- read_csv("data/experiment_table.csv", show_col_types = FALSE) |>
  column_to_rownames("ID")

# Factors; set Untreated as the reference so LFCs read as Dex-vs-Untreated
coldata$treatment <- relevel(factor(coldata$treatment), ref = "Untreated")
coldata$cellLine  <- factor(coldata$cellLine)

# --- Align matrix columns to metadata rows (fail loudly if mismatched) -----
stopifnot(setequal(colnames(count_mat), rownames(coldata)))
count_mat <- count_mat[, rownames(coldata)]
stopifnot(identical(colnames(count_mat), rownames(coldata)))

dim(count_mat)
coldata
```

# Build the DESeqDataSet

```{r}
#| label: build-dds

dds <- DESeqDataSetFromMatrix(
  countData = count_mat,
  colData   = coldata,
  design    = ~ cellLine + treatment
)

# Pre-filter: keep genes with >= 10 counts in at least `smallest group` samples.
# Smallest group here = number of samples per treatment level (4).
smallest_group <- min(table(coldata$treatment))
keep <- rowSums(counts(dds) >= 10) >= smallest_group
message(sprintf("Genes retained after filtering: %d of %d",
                sum(keep), length(keep)))
dds <- dds[keep, ]
```

# Run DESeq2

```{r}
#| label: run-deseq

dds <- DESeq(dds)
resultsNames(dds)
```

# Quality control / exploratory analysis

Variance-stabilising transform (VST) for visualisation. Because the modelled
`treatment` effect is small relative to library-size variation, VST-normalised
values give a fairer picture than raw counts.

```{r}
#| label: vst

vsd <- vst(dds, blind = FALSE)
```

## PCA

```{r}
#| label: pca
#| fig-cap: "PCA of VST-transformed counts, coloured by treatment and shaped by cell line."

pca_df <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
pct <- round(100 * attr(pca_df, "percentVar"))

p_pca <- ggplot(pca_df, aes(PC1, PC2, colour = treatment, shape = cellLine)) +
  geom_point(size = 4) +
  labs(x = sprintf("PC1: %d%% variance", pct[1]),
       y = sprintf("PC2: %d%% variance", pct[2]),
       title = "PCA — VST counts") +
  theme_bw()

p_pca
ggsave("figures/pca.png", p_pca, width = 7, height = 5, dpi = 300)
```

## Sample-to-sample distances

```{r}
#| label: sample-dist
#| fig-cap: "Euclidean distances between samples on VST-transformed data."

samp_dist <- dist(t(assay(vsd)))
dist_mat  <- as.matrix(samp_dist)
rownames(dist_mat) <- paste(vsd$cellLine, vsd$treatment, sep = " | ")
colnames(dist_mat) <- NULL
cols <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

png("figures/sample_distance_heatmap.png", width = 7, height = 6,
    units = "in", res = 300)
pheatmap(dist_mat,
         clustering_distance_rows = samp_dist,
         clustering_distance_cols = samp_dist,
         col = cols,
         main = "Sample-to-sample distances")
dev.off()

pheatmap(dist_mat,
         clustering_distance_rows = samp_dist,
         clustering_distance_cols = samp_dist,
         col = cols,
         main = "Sample-to-sample distances")
```

# Differential expression results

Contrast of interest: **Dexamethasone vs Untreated**. We apply `apeglm` LFC
shrinkage for stable effect-size ranking (volcano / gene shortlists), while
keeping the unshrunken table for reference.

```{r}
#| label: results

coef_name <- "treatment_Dexamethasone_vs_Untreated"
stopifnot(coef_name %in% resultsNames(dds))

# Unshrunken results (for p-values / summary)
res <- results(dds, name = coef_name, alpha = alpha_padj)
summary(res)

# Shrunken LFCs (for ranking & visualisation)
res_shrink <- lfcShrink(dds, coef = coef_name, type = "apeglm")
```

```{r}
#| label: results-table

res_df <- res_shrink |>
  as.data.frame() |>
  rownames_to_column("gene_id") |>
  # bring in unshrunken stat + pvalue for completeness
  left_join(
    res |> as.data.frame() |> rownames_to_column("gene_id") |>
      select(gene_id, stat, pvalue_raw = pvalue),
    by = "gene_id"
  ) |>
  arrange(padj)

head(res_df, 10)
```

## Gene-symbol annotation

Map Ensembl gene IDs to HGNC symbols using `org.Hs.eg.db` — fully offline (no
biomaRt / network call). A `label` column carries the symbol when available and
falls back to the Ensembl ID otherwise; it is used for the plot labels below.

```{r}
#| label: annotate

if (requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  suppressPackageStartupMessages(library(org.Hs.eg.db))
  res_df$symbol <- mapIds(org.Hs.eg.db,
                          keys = res_df$gene_id,
                          column = "SYMBOL",
                          keytype = "ENSEMBL",
                          multiVals = "first")
  message(sprintf("Annotated %d of %d genes with HGNC symbols.",
                  sum(!is.na(res_df$symbol)), nrow(res_df)))
} else {
  res_df$symbol <- NA_character_
  message("org.Hs.eg.db not installed - keeping Ensembl IDs only.")
}

res_df <- res_df |>
  mutate(label = ifelse(is.na(symbol) | symbol == "", gene_id, symbol)) |>
  relocate(symbol, label, .after = gene_id)

head(res_df, 10)
```

# Diagnostics

```{r}
#| label: ma-plot
#| fig-cap: "MA plot (shrunken LFCs)."

png("figures/ma_plot.png", width = 7, height = 5, units = "in", res = 300)
plotMA(res_shrink, ylim = c(-5, 5), main = "MA plot — Dex vs Untreated")
dev.off()
plotMA(res_shrink, ylim = c(-5, 5), main = "MA plot — Dex vs Untreated")
```

```{r}
#| label: dispersion
#| fig-cap: "Dispersion estimates."

png("figures/dispersion.png", width = 7, height = 5, units = "in", res = 300)
plotDispEsts(dds, main = "Dispersion estimates")
dev.off()
plotDispEsts(dds, main = "Dispersion estimates")
```

```{r}
#| label: pval-hist
#| fig-cap: "Raw p-value distribution (independent-filtering diagnostic)."

hist(res$pvalue, breaks = 50, col = "grey", border = "white",
     xlab = "raw p-value", main = "P-value histogram")
```

# Top-gene visualisations

## Volcano plot

```{r}
#| label: volcano
#| fig-cap: "Volcano plot of shrunken LFC vs -log10 adjusted p-value."

volc <- res_df |>
  filter(!is.na(padj)) |>
  mutate(sig = case_when(
    padj < alpha_padj & log2FoldChange >=  lfc_thresh ~ "Up",
    padj < alpha_padj & log2FoldChange <= -lfc_thresh ~ "Down",
    TRUE ~ "NS"
  ))

top_lab <- volc |> filter(sig != "NS") |> slice_min(padj, n = 15)

p_volc <- ggplot(volc, aes(log2FoldChange, -log10(padj), colour = sig)) +
  geom_point(alpha = 0.6, size = 1.4) +
  scale_colour_manual(values = c(Up = "#c0392b", Down = "#2471a3", NS = "grey70")) +
  geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dashed", colour = "grey50") +
  geom_hline(yintercept = -log10(alpha_padj), linetype = "dashed", colour = "grey50") +
  ggrepel::geom_text_repel(data = top_lab, aes(label = label),
                           size = 2.6, max.overlaps = 20, show.legend = FALSE) +
  labs(x = "log2 fold change (Dex vs Untreated)",
       y = "-log10 adjusted p-value", colour = NULL,
       title = "Volcano — Dexamethasone vs Untreated") +
  theme_bw()

p_volc
ggsave("figures/volcano.png", p_volc, width = 7, height = 6, dpi = 300)
```

## Heatmap of top differentially expressed genes

```{r}
#| label: top-heatmap
#| fig-cap: "Top 30 DE genes by adjusted p-value (VST, row z-scored)."

top_genes <- res_df |> filter(!is.na(padj)) |> slice_min(padj, n = 30) |> pull(gene_id)

mat <- assay(vsd)[top_genes, ]
mat <- t(scale(t(mat)))  # row z-score
rownames(mat) <- res_df$label[match(top_genes, res_df$gene_id)]  # symbol labels

ann_col <- as.data.frame(colData(dds)[, c("treatment", "cellLine")])

png("figures/top_genes_heatmap.png", width = 8, height = 8, units = "in", res = 300)
pheatmap(mat, annotation_col = ann_col,
         show_colnames = FALSE, fontsize_row = 7,
         main = "Top 30 DE genes (VST z-score)")
dev.off()

pheatmap(mat, annotation_col = ann_col,
         show_colnames = FALSE, fontsize_row = 7,
         main = "Top 30 DE genes (VST z-score)")
```

# Export outputs

```{r}
#| label: export

# Full results
write_csv(res_df, "results/deseq2_results_full.csv")

# Significant subset
sig_df <- res_df |> filter(!is.na(padj), padj < alpha_padj) |> arrange(padj)
write_csv(sig_df, "results/deseq2_results_significant.csv")

# Normalised counts
norm_counts <- counts(dds, normalized = TRUE) |>
  as.data.frame() |>
  rownames_to_column("gene_id")
write_csv(norm_counts, "results/normalized_counts.csv")

message(sprintf("Significant genes (padj < %.2f): %d", alpha_padj, nrow(sig_df)))
message("Wrote: results/deseq2_results_full.csv, results/deseq2_results_significant.csv, results/normalized_counts.csv")
```

# Session info

```{r}
#| label: session-info

sessionInfo()
```
