---
title: "Dexamethasone response in airway smooth muscle cells"
subtitle: "Bulk RNA-seq differential expression with DESeq2"
author: "Tiphaine Cayol"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 7
    fig-height: 5
execute:
  warning: false
  message: false
---

## Overview

Differential expression analysis of the *airway* dataset: 4 human airway smooth muscle
cell lines, each with a paired untreated and dexamethasone-treated sample (Himes et al.
2014, PMID 24926665).

The design is **paired by donor cell line** — each line supplies its own control — so the
model blocks on `cellLine` and tests `treatment`:

$$\text{design} = \sim \text{cellLine} + \text{treatment}$$

```{r}
#| label: setup

library(DESeq2)
library(tidyverse)
library(pheatmap)
library(ggrepel)
library(RColorBrewer)

set.seed(42)

knitr::opts_chunk$set(dev = "png", dpi = 150)

# Resolve the repo root whether this is rendered from the repo root or from analysis/.
# Avoids a dependency on {here} while keeping paths correct either way.
repo_root <- if (file.exists("data/experiment_table.csv")) "." else ".."
stopifnot(file.exists(file.path(repo_root, "data", "experiment_table.csv")))

fig_dir <- file.path(repo_root, "results", "figures")
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

ALPHA <- 0.05  # FDR threshold, used consistently throughout
```

## Load inputs

The count matrix carries a `transcript_id(s)` column alongside `gene_id`; both are dropped
from the numeric matrix, with `gene_id` retained as rownames.

```{r}
#| label: load-data

counts_raw <- read_tsv(
  file.path(repo_root, "data", "rsem.merged.gene_counts.tsv"),
  show_col_types = FALSE
)

coldata <- read_csv(
  file.path(repo_root, "data", "experiment_table.csv"),
  show_col_types = FALSE
)

gene_ids <- counts_raw$gene_id
stopifnot(!anyDuplicated(gene_ids))  # rownames must be unique

counts <- counts_raw |>
  select(-gene_id, -`transcript_id(s)`) |>
  as.matrix()
rownames(counts) <- gene_ids

cat(sprintf("Loaded %d genes x %d samples\n", nrow(counts), ncol(counts)))
knitr::kable(coldata, caption = "Sample sheet")
```

## Round non-integer counts

RSEM apportions multi-mapping reads probabilistically, so counts are **estimates, not
integers** — `r format(sum(apply(counts, 1, function(x) any(x != floor(x)))), big.mark = ",")`
genes carry fractional values here. DESeq2 requires integers and errors otherwise.

`round()` is used deliberately rather than `floor()`, `trunc()`, or `as.integer()`: those
truncate downward and would bias low-count genes toward zero.

```{r}
#| label: round-counts

n_fractional <- sum(apply(counts, 1, function(x) any(x != floor(x))))
zeros_before <- sum(rowSums(counts) == 0)

counts <- round(counts)
storage.mode(counts) <- "integer"  # DESeq2 wants an integer matrix

zeros_after <- sum(rowSums(counts) == 0)

cat(sprintf(
  "Genes with fractional counts: %s\nAll-zero genes before rounding: %s\nAll-zero genes after rounding:  %s\n",
  format(n_fractional, big.mark = ","),
  format(zeros_before, big.mark = ","),
  format(zeros_after, big.mark = ",")
))
```

Note the all-zero count *rises* after rounding — genes whose every value sat below 0.5 collapse
to zero. This is why filtering must happen **after** rounding, never before.

## Align samples to the sample sheet

Column order happens to match the sample sheet in this file, but relying on that is how
silently-wrong results happen. Subset by name and assert.

```{r}
#| label: align-samples

counts <- counts[, coldata$ID]
stopifnot(identical(colnames(counts), coldata$ID))

cat("Counts columns and sample sheet rows correspond by name.\n")
```

## Factors and reference level

R orders factor levels alphabetically, which would make `Dexamethasone` the reference and
**invert the sign of every log2 fold change**. Setting `Untreated` explicitly means a positive
LFC reads as "induced by dexamethasone".

```{r}
#| label: factors

# read_csv returns a tibble, which has no real rownames. DESeq2 matches colData rows to
# count columns by rowname, and pheatmap needs them for annotations, so make it a
# base data.frame keyed by sample ID.
coldata <- as.data.frame(coldata)
rownames(coldata) <- coldata$ID

coldata$treatment <- relevel(factor(coldata$treatment), ref = "Untreated")
coldata$cellLine  <- factor(coldata$cellLine)

cat("treatment levels:", paste(levels(coldata$treatment), collapse = " -> "), "\n")
cat("cellLine levels: ", paste(levels(coldata$cellLine), collapse = ", "), "\n")

stopifnot(levels(coldata$treatment)[1] == "Untreated")
```

The first level printed above is the reference. `Untreated` must appear first.

## Pre-filter low-count genes

Nearly half of all genes are zero across every sample. Keeping genes with at least 10 counts
in at least 4 samples (4 = the smallest group size) drops uninformative genes, which improves
the dispersion fit and reduces the multiple-testing burden.

```{r}
#| label: prefilter

n_before <- nrow(counts)
keep <- rowSums(counts >= 10) >= 4
counts <- counts[keep, ]

cat(sprintf(
  "Genes before filtering: %s\nGenes after filtering:  %s (%.1f%% retained)\n",
  format(n_before, big.mark = ","),
  format(nrow(counts), big.mark = ","),
  100 * nrow(counts) / n_before
))
```

## Build the DESeqDataSet and fit

```{r}
#| label: run-deseq

dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = coldata,
  design    = ~ cellLine + treatment
)

dds <- DESeq(dds)

resultsNames(dds)
```

`treatment_Dexamethasone_vs_Untreated` is the coefficient of interest; the `cellLine_*` terms
absorb donor-to-donor baseline differences and are not interpreted.

## Quality control

### Library sizes

```{r}
#| label: qc-libsize

libsize <- tibble(
  ID       = colnames(dds),
  total    = colSums(counts(dds)),
  treatment = colData(dds)$treatment,
  cellLine  = colData(dds)$cellLine
)

p_lib <- ggplot(libsize, aes(x = ID, y = total / 1e6, fill = treatment)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Assigned reads (millions)", title = "Library sizes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_lib
ggsave(file.path(fig_dir, "library_sizes.png"), p_lib, width = 7, height = 4.5, dpi = 150)
```

Library sizes span roughly a two-fold range, which is why raw counts are never compared
directly — DESeq2's size factors handle this.

### PCA

Variance-stabilised counts, computed blind to the design so QC stays independent of the
model being tested.

```{r}
#| label: qc-pca

vsd <- vst(dds, blind = TRUE)

pca_data <- plotPCA(vsd, intgroup = c("treatment", "cellLine"), returnData = TRUE)
pct_var <- round(100 * attr(pca_data, "percentVar"))

p_pca <- ggplot(pca_data, aes(PC1, PC2, colour = treatment, shape = cellLine)) +
  geom_point(size = 4, alpha = 0.9) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    x = paste0("PC1: ", pct_var[1], "% variance"),
    y = paste0("PC2: ", pct_var[2], "% variance"),
    title = "PCA of variance-stabilised counts"
  ) +
  theme_minimal()

p_pca
ggsave(file.path(fig_dir, "pca.png"), p_pca, width = 7, height = 5, dpi = 150)
```

In a well-behaved paired design one axis separates treatment and another separates donor.
Because donor differences are often the larger effect, treatment may well sit on PC2 rather
than PC1 — that is expected here, not a problem, and is exactly what blocking on `cellLine`
accounts for.

### Sample-to-sample distances

```{r}
#| label: qc-distance
#| fig-height: 5.5

sample_dists <- dist(t(assay(vsd)))
dist_matrix <- as.matrix(sample_dists)
rownames(dist_matrix) <- paste(colData(vsd)$cellLine, colData(vsd)$treatment, sep = " | ")
colnames(dist_matrix) <- colnames(vsd)

dist_args <- list(
  mat = dist_matrix,
  clustering_distance_rows = sample_dists,
  clustering_distance_cols = sample_dists,
  color = colorRampPalette(rev(brewer.pal(9, "Blues")))(255),
  main = "Sample-to-sample distances"
)

do.call(pheatmap, dist_args)

# pheatmap writes files via its own `filename` argument; a manual png()/dev.off()
# wrapper does not capture its grid output reliably.
invisible(do.call(pheatmap, c(
  dist_args,
  list(filename = file.path(fig_dir, "sample_distances.png"), width = 7, height = 5.5)
)))
```

Samples are expected to pair up by cell line, reflecting donor as the dominant structure.

## Differential expression results

Two versions are kept deliberately: **unshrunk** effect sizes for the statistics, and
**shrunk** effect sizes for ranking and plotting. Shrinkage pulls in the noisy fold changes
of low-count genes without touching p-values.

```{r}
#| label: results

res <- results(
  dds,
  contrast = c("treatment", "Dexamethasone", "Untreated"),
  alpha = ALPHA
)

# apeglm is the preferred shrinkage estimator; degrade gracefully if it is absent
# so a missing optional package cannot fail the render.
shrink_type <- if (requireNamespace("apeglm", quietly = TRUE)) "apeglm" else "normal"
if (shrink_type == "normal") {
  message("apeglm not installed - falling back to type = 'normal' shrinkage.")
}

res_shrunk <- lfcShrink(
  dds,
  coef = "treatment_Dexamethasone_vs_Untreated",
  type = shrink_type
)

summary(res)
```

```{r}
#| label: results-counts

n_sig  <- sum(res$padj < ALPHA, na.rm = TRUE)
n_up   <- sum(res$padj < ALPHA & res$log2FoldChange > 0, na.rm = TRUE)
n_down <- sum(res$padj < ALPHA & res$log2FoldChange < 0, na.rm = TRUE)

cat(sprintf(
  "At FDR < %.2f: %s genes differentially expressed (%s up, %s down) out of %s tested.\nShrinkage estimator: %s\n",
  ALPHA,
  format(n_sig, big.mark = ","), format(n_up, big.mark = ","),
  format(n_down, big.mark = ","), format(sum(!is.na(res$padj)), big.mark = ","),
  shrink_type
))
```

### Annotate with gene symbols

The matrix carries Ensembl IDs only. `org.Hs.eg.db` maps them offline, keeping the report
reproducible; if it is unavailable the analysis proceeds on Ensembl IDs alone.

```{r}
#| label: annotate

res_df <- as.data.frame(res) |>
  rownames_to_column("gene_id") |>
  mutate(log2FoldChange_shrunk = res_shrunk$log2FoldChange[match(gene_id, rownames(res_shrunk))])

if (requireNamespace("org.Hs.eg.db", quietly = TRUE)) {
  res_df$symbol <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys      = res_df$gene_id,
    keytype   = "ENSEMBL",
    column    = "SYMBOL",
    multiVals = "first"
  )
  cat(sprintf("Mapped symbols for %s of %s genes.\n",
              format(sum(!is.na(res_df$symbol)), big.mark = ","),
              format(nrow(res_df), big.mark = ",")))
} else {
  res_df$symbol <- NA_character_
  message("org.Hs.eg.db not installed - results reported with Ensembl IDs only.")
}

# Label preferring symbol, falling back to Ensembl ID
res_df <- res_df |>
  mutate(label = coalesce(symbol, gene_id)) |>
  arrange(padj)
```

### Top differentially expressed genes

```{r}
#| label: top-genes

res_df |>
  slice_head(n = 25) |>
  select(gene_id, symbol, baseMean, log2FoldChange, log2FoldChange_shrunk, pvalue, padj) |>
  mutate(
    baseMean = round(baseMean, 1),
    across(starts_with("log2FoldChange"), \(x) round(x, 3)),
    # p-values span many orders of magnitude; scientific notation keeps them readable
    across(c(pvalue, padj), \(x) format(x, scientific = TRUE, digits = 3))
  ) |>
  knitr::kable(caption = "Top 25 genes by adjusted p-value")
```

### Sanity check against published markers

Dexamethasone is a glucocorticoid; these four genes are well-established responders in this
system and should all be **induced** (positive LFC). Matching is done by Ensembl ID so the
check holds even without the annotation package.

**If any fold change here is negative, the reference level is wrong** — that is the failure
this check exists to catch.

```{r}
#| label: sanity-check

markers <- tribble(
  ~gene_id,          ~expected_symbol,
  "ENSG00000120129", "DUSP1",
  "ENSG00000163884", "KLF15",
  "ENSG00000179094", "PER1",
  "ENSG00000103196", "CRISPLD2"
)

marker_res <- markers |>
  left_join(res_df, by = "gene_id") |>
  select(gene_id, expected_symbol, mapped_symbol = symbol,
         baseMean, log2FoldChange, padj)

marker_res |>
  mutate(
    baseMean = round(baseMean, 1),
    log2FoldChange = round(log2FoldChange, 3),
    padj = format(padj, scientific = TRUE, digits = 3)
  ) |>
  knitr::kable(caption = "Known dexamethasone-responsive genes")

if (all(marker_res$log2FoldChange > 0, na.rm = TRUE)) {
  cat("PASS: all marker genes show positive log2 fold change (induced), as expected.\n")
} else {
  warning("Marker genes show unexpected direction - check the reference level and contrast.")
}
```

## Figures

### MA plot

```{r}
#| label: fig-ma

p_ma <- res_df |>
  filter(!is.na(padj)) |>
  mutate(significant = padj < ALPHA) |>
  ggplot(aes(x = baseMean, y = log2FoldChange_shrunk, colour = significant)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey30") +
  scale_x_log10() +
  scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#d95f02"),
                      labels = c(`FALSE` = "n.s.", `TRUE` = paste0("FDR < ", ALPHA))) +
  labs(x = "Mean of normalised counts", y = "log2 fold change (shrunk)",
       colour = NULL, title = "MA plot: Dexamethasone vs Untreated") +
  theme_minimal()

p_ma
ggsave(file.path(fig_dir, "ma_plot.png"), p_ma, width = 7, height = 5, dpi = 150)
```

### Volcano plot

```{r}
#| label: fig-volcano

volcano_data <- res_df |>
  filter(!is.na(padj)) |>
  mutate(
    significant = padj < ALPHA & abs(log2FoldChange_shrunk) > 1,
    neg_log10_padj = -log10(padj)
  )

p_volcano <- ggplot(volcano_data,
                    aes(x = log2FoldChange_shrunk, y = neg_log10_padj, colour = significant)) +
  geom_point(alpha = 0.4, size = 0.7) +
  geom_hline(yintercept = -log10(ALPHA), linetype = "dashed", colour = "grey30") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey30") +
  geom_text_repel(
    data = slice_head(arrange(volcano_data, padj), n = 20),
    aes(label = label), size = 3, max.overlaps = 20, show.legend = FALSE
  ) +
  scale_colour_manual(values = c(`FALSE` = "grey70", `TRUE` = "#7570b3"),
                      labels = c(`FALSE` = "n.s.", `TRUE` = paste0("FDR < ", ALPHA, " & |LFC| > 1"))) +
  labs(x = "log2 fold change (shrunk)", y = expression(-log[10]~adjusted~italic(p)),
       colour = NULL, title = "Volcano plot: Dexamethasone vs Untreated") +
  theme_minimal()

p_volcano
ggsave(file.path(fig_dir, "volcano_plot.png"), p_volcano, width = 8, height = 6, dpi = 150)
```

### Heatmap of top differentially expressed genes

Variance-stabilised counts, centred per gene so the colour shows deviation from each gene's
mean rather than absolute expression.

```{r}
#| label: fig-heatmap
#| fig-height: 8

top_genes <- res_df |>
  filter(!is.na(padj)) |>
  slice_head(n = 30)

mat <- assay(vsd)[top_genes$gene_id, ]
mat <- mat - rowMeans(mat)
rownames(mat) <- top_genes$label

annotation_col <- as.data.frame(colData(vsd)[, c("treatment", "cellLine")])

heatmap_args <- list(
  mat = mat,
  annotation_col = annotation_col,
  show_colnames = TRUE,
  cluster_cols = TRUE,
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(255),
  main = "Top 30 DE genes (centred VST)"
)

do.call(pheatmap, heatmap_args)

invisible(do.call(pheatmap, c(
  heatmap_args,
  list(filename = file.path(fig_dir, "heatmap_top_de_genes.png"), width = 8, height = 8)
)))
```

## Export results

```{r}
#| label: export

results_path <- file.path(repo_root, "results", "deseq2_results.csv")

res_df |>
  select(gene_id, symbol, baseMean, log2FoldChange, log2FoldChange_shrunk,
         lfcSE, stat, pvalue, padj) |>
  write_csv(results_path)

cat(sprintf("Wrote %s rows to %s\n",
            format(nrow(res_df), big.mark = ","), results_path))
cat("Figures written to", fig_dir, "\n")
```

## Session info

```{r}
#| label: session-info

sessionInfo()
```
